---
title: "Translate contrasts into feature effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Translate contrasts into feature effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Backprojection expresses a factor-activity contrast as a difference in the
original feature units. The example compares brain tissues with all other
tissues in the GTEx z-score matrix that ships with flashier.

```{r setup}
library(flashier.utils)
data("gtex", package = "flashier")
fit <- fit_ebmf(gtex, sample_side = "columns", max_factors = 3,
                backfit = TRUE, seed = 1, verbose = 0,
                feature_scale = "association_z_score")
view <- standardize_factors(fit)
```

Rows pair a SNP with a gene, and columns are 44 tissues. The entries are
association z-scores, so feature estimates below are differences in mean
z-score between tissue groups.

## Compare groups

The groups come from tissue names alone: the ten tissues whose names start with
`Brain_` against the other 34. Tissues are not independent donors, so the
contrast is descriptive and has no donor-level replication or p-value.

```{r contrast}
is_brain <- startsWith(colnames(gtex), "Brain_")
metadata <- data.frame(
  sample = colnames(gtex), group = ifelse(is_brain, "Brain", "Other")
)
table(metadata$group)
difference <- compare_factor_groups(view, metadata, "group",
                                    c("Brain", "Other"))
difference$factor_contrast[, c("factor", "estimate")]
head(difference$feature_contrast[, c("feature", "estimate")])
```

`compare_factor_groups()` returns both factor and feature contrasts.
The direction here is Brain minus Other, with each tissue weighted equally.
F2 has the largest factor contrast, in agreement with its brain-specific
activity in [Fit and interpret factors](first-fit.html).
Feature estimates retain the input matrix's units: here, z-score differences.

## Check the reconstruction

Compare the represented difference with the difference calculated directly
from the same samples and weights.

```{r validation}
observed <- data.frame(
  feature = rownames(gtex),
  estimate = rowMeans(gtex[, is_brain]) - rowMeans(gtex[, !is_brain]),
  contrast = "Brain - Other"
)
validate_backprojection(difference$feature_contrast, observed,
                        feature_scale = "association_z_score")$metrics
```

The factorization can omit residual patterns, so the differences need not
agree exactly. Here the three-factor reconstruction tracks the observed
difference closely across the 1000 SNP-gene pairs, but the slope below one
shows that it shrinks the observed differences on average.
Prediction R-squared can be negative if the reconstruction performs worse than
a constant prediction.

## Project coefficients from a model

For detached factor coefficients, supply the representation used to fit the
activity model. This example projects a one-unit increase in every factor.
The coefficients are placeholders, not estimates from a model.

```{r coefficients}
beta <- setNames(rep(1, ncol(view$activity)), colnames(view$activity))
projected <- backproject_contrast(view, beta, input = "coefficients",
                                  factor_basis = view$manifest)
head(projected$effects)
```

Coefficients must match the factor scale, sign and order. Point coefficients
alone produce point feature effects without p-values.
See [Understand uncertainty](uncertainty.html) for coefficient draws and covariance.

## Sources

The GTEx data ship with flashier (Jason Willwerscheid, Peter Carbonetto, Wei
Wang and Matthew Stephens) and derive from the GTEx project; see `?flashier::gtex`.

Reconstruction follows the FlashierDGE and ImmGenT workflows by David Zemmour
and the Zemmour Lab. The implementation uses independently written linear algebra.
[Original authors and sources](credits.html) lists the workflow credits.
