---
title: "Translate contrasts into feature effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Translate contrasts into feature effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Backprojection expresses a factor-activity contrast as a difference in the
original feature units. The example compares two groups in a simulated matrix.

```{r setup, echo=FALSE}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48), 24, 2), matrix(rnorm(32), 16, 2)) +
  matrix(rnorm(384, sd = 0.2), 24, 16)
dimnames(X) <- list(paste0("s", 1:24), paste0("g", 1:16))
fit <- fit_ebmf(X, sample_side = "rows", seed = 1, max_factors = 2,
                backfit = TRUE, nullcheck = FALSE, verbose = 0,
                feature_scale = "simulated_intensity")
view <- standardize_factors(fit)
```

## Compare groups

```{r contrast}
metadata <- data.frame(
  sample = rownames(X), group = rep(c("A", "B"), each = 12),
  donor = rep(paste0("d", 1:12), each = 2)
)
difference <- compare_factor_groups(view, metadata, "group", c("A", "B"),
                                    unit_col = "donor")
head(difference$feature_contrast)
```

`compare_factor_groups()` returns both factor and feature contrasts.
The direction here is A minus B. Feature estimates retain the input matrix's
units. For arcsinh input, the estimates are differences in arcsinh units.

## Check the reconstruction

Compare the represented difference with the difference calculated directly
from the same samples and weights.

```{r validation}
observed <- data.frame(
  feature = colnames(X),
  estimate = colMeans(X[1:12, ]) - colMeans(X[13:24, ]),
  contrast = "A - B"
)
validate_backprojection(difference$feature_contrast, observed,
                        feature_scale = "simulated_intensity")$metrics
```

The factorization can omit residual patterns, so the differences need not
agree exactly. Prediction R-squared can be negative if the reconstruction
performs worse than a constant prediction.

## Project coefficients from a model

For detached factor coefficients, supply the representation used to fit the
activity model. This example projects a one-unit increase in every factor.

```{r coefficients}
beta <- setNames(rep(1, ncol(view$activity)), colnames(view$activity))
projected <- backproject_contrast(view, beta, input = "coefficients",
                                  factor_basis = view$manifest)
head(projected$effects)
```

Coefficients must match the factor scale, sign and order. Point coefficients
alone produce point feature effects without p-values.
See [Understand uncertainty](uncertainty.html) for coefficient draws and covariance.

## Sources

Reconstruction follows the FlashierDGE and ImmGenT workflows by David Zemmour
and the Zemmour Lab. The implementation uses independently written linear algebra.
[Original authors and sources](credits.html) lists the workflow credits.
