---
title: "Compare groups and test associations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare groups and test associations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(flashier.utils)
library(Matrix)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5)
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache_file <- file.path(pancreas_dir,"pancreas-cells.rds")
have_pancreas <- nzchar(pancreas_dir) && file.exists(cache_file)
knitr::opts_chunk$set(eval=have_pancreas)
```

Use group summaries to describe activity, or fit a regression to test an
association. In either case, metadata must identify samples by name.

The example uses **real pancreas cells** from
[Peter Carbonetto's single-cell Jamboree](https://stephenslab.github.io/single-cell-jamboree/pancreas.html)
(pinned revision 6cf9aa72): human pancreas cells from the cross-technology integration benchmark of
Luecken et al. (2022) as assembled there; the CEL-Seq2 cells are from Muraro et
al. (2016), and the Jamboree page links the original studies of the other
technologies. A local cache holds up to 40 cells per technology and cell type
(CEL-Seq2, SMART-Seq2, Fluidigm C1 and inDrop3; 896 cells). It is built outside
package builds by the repository script `.github/scripts/prepare-pancreas.R`, so
data chunks run only when `FLASHIER_UTILS_PANCREAS_DIR` points to that cache, and
nothing is downloaded. The cells are not redistributed with the package. Native
EBMF fitting credits Jason Willwerscheid, Peter Carbonetto, Wei Wang and
Matthew Stephens.

## Fit factors and declare metadata

Scale each cell to the CEL-Seq2 reference library mean, apply `log1p`, keep the
1,000 genes with the largest variance and fit eight non-negative factors with
`fit_ebmf()`. These are fitted here; they are not the published Jamboree factors.
The gene selection uses every cell, so it is a global choice.

```{r fit}
cache <- readRDS(cache_file)
metadata <- cache$metadata
metadata$sample <- as.character(metadata$id)
rownames(metadata) <- metadata$sample
metadata$celltype <- droplevels(factor(metadata$celltype))
metadata$tech <- droplevels(factor(metadata$tech))
counts <- cache$counts
stopifnot(identical(rownames(counts), metadata$sample))
Y <- Diagonal(x = cache$reference_library_mean / rowSums(counts)) %*% counts
Y <- log1p(Y)
dimnames(Y) <- dimnames(counts)
variance <- colMeans(Y^2) - colMeans(Y)^2
genes <- names(sort(variance, decreasing = TRUE))[1:1000]
X <- as.matrix(Y[, genes])
fit <- fit_ebmf(X, sample_side = "rows", ebnm_fn = ebnm::ebnm_point_exponential,
                max_factors = 8, backfit = TRUE, nullcheck = FALSE, seed = 1,
                verbose = 0,
                feature_scale = "shifted_log_reference_library_normalized_measurements")
view <- standardize_factors(fit)
dim(view$activity)
```


The cache has no donor identifiers. The two metadata columns that can define
groups are the cell type and the measurement technology. Cells from one
technology share a platform and batch, so they are not independent units.

## Describe group differences

```{r summaries}
summary <- summarize_factor_activity(view, metadata, by = "celltype", unit_col = "tech")
summary[summary$factor %in% c("F4", "F5"), c("celltype", "factor", "n", "n_units", "mean")]
difference <- compare_factor_groups(view, metadata, "celltype", c("beta", "alpha"),
                                    unit_col = "tech")
difference$factor_contrast[, c("factor", "contrast", "estimate")]
```

The contrast is beta minus alpha. These summaries have no p-values.
Means weight cells equally by default. `unit_col` counts the units in each group
but does not balance their contributions: here the units are the four
technologies, and Fluidigm C1 contributes fewer acinar, delta, ductal and gamma
cells than the other technologies, so group sizes are unequal.

## Test an association

`engine = "lm"` and `engine = "limma"` require **one record per independent
unit**, declared through `unit_col`. This cache has no donor or other
biological-unit identifier, and cells from one donor or one technology are
correlated, so declaring each cell as an independent unit would be an assumption
the data cannot support. This vignette therefore does not run those engines on
cells. With independent donors, aggregate or select one record per donor and use
`lm`; see `?test_factors`.

What the data do support is a model that accounts for technology as a grouping
factor. Install variancePartition to use `dream`. Declare the technology as the
grouping unit with a random intercept and retain every cell. The reference
level of `celltype` is alpha, so the `celltypebeta` term estimates beta minus alpha.

```{r repeated}
metadata$celltype <- relevel(metadata$celltype, ref = "alpha")
if (requireNamespace("variancePartition", quietly = TRUE)) {
  repeated <- test_factors(
    view, metadata, ~celltype + (1 | tech), engine = "dream",
    unit_col = "tech", independence = "repeated",
    control = list(dream = list(BPPARAM = BiocParallel::SerialParam()))
  )
  beta_alpha <- repeated$table$term == "celltypebeta"
  print(repeated$table[beta_alpha, c("factor", "term", "estimate", "std.error", "adj.p.value", "status")])
  repeated$diagnostics[c("warnings", "errors")]
}
```

The `celltypebeta` estimates equal the descriptive beta-minus-alpha contrasts
above, because the cell-type groups are balanced across technologies in this
comparison. The clear differences are `F4` (negative, the alpha factor) and `F5`
(positive, the beta factor); `F6`, one of the technology factors, has a small
positive estimate, and the others are near zero. Check `status` and
`diagnostics` before interpreting a test. All tests condition on the fitted activities and selected number of factors.
They do not propagate uncertainty from learning the factorization. A technology
random intercept absorbs shifts shared by all cells of a technology, but it
does not model donor-level correlation among cells, and it is estimated from
only four technologies. Adjusted p-values near zero reflect 320 cells in the
contrast, not independent donors, so treat them as descriptive screening rather
than donor-level inference.

To express a contrast in feature units, see
[Translate contrasts into feature effects](backprojection.html).

## Sources

The data are from Peter Carbonetto's
[single-cell Jamboree](https://github.com/stephenslab/single-cell-jamboree)
(Muraro et al., 2016; Luecken et al., 2022, as assembled there).
Group summaries follow workflows by David Zemmour and the Zemmour Lab.
Tests call Gabriel Hoffman's dream (variancePartition) or, for independent
units, R Core Team's `lm` and Gordon Smyth and contributors' limma.
[Original authors and sources](credits.html) lists citations.
