---
title: "Descriptive and inferential contrasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Descriptive and inferential contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48),24,2),matrix(rnorm(32),16,2)) + matrix(rnorm(384,sd=0.2),24,16)
dimnames(X) <- list(paste0("s",1:24),paste0("g",1:16))
fit <- fit_ebmf(X,"rows",seed=1,max_factors=2,verbose=0,backfit=TRUE,nullcheck=FALSE,feature_scale="arcsinh_intensity")
view <- standardize_factors(fit)
metadata <- data.frame(sample=rownames(X),donor=rep(paste0("d",1:12),each=2),group=rep(c("A","B"),each=12),visit=rep(0:1,12))
```

Cells and visits are observations; donors are biological units. Descriptive group means provide no p-values. Factor-response tests condition on the estimated activities and selected rank.
```{r}
summarize_factor_activity(view,metadata,by="group",unit_col="donor")
compare_factor_groups(view,metadata,"group",c("B","A"),unit_col="donor")
independent <- metadata[seq(1,24,2),]
# One observation per donor; explicit sample selection creates a native fit.
indfit <- fit_ebmf(X[independent$sample,],"rows",seed=8,max_factors=2,verbose=0,nullcheck=FALSE,feature_scale="arcsinh_intensity")
result <- test_factors(indfit,independent,~group,"lm",unit_col="donor",independence="independent")
result$table
```
Repeated records must use the declared donor term. Native dream diagnostics, including errors, are preserved.
```{r}
if (requireNamespace("variancePartition",quietly=TRUE)) {
  repeated <- test_factors(view,metadata,~visit+(1|donor),"dream",unit_col="donor",independence="repeated",control=list(dream=list(BPPARAM=BiocParallel::SerialParam())))
  repeated$table
  repeated$diagnostics
}
```
Limma moderation is optional and sensitive with few factors. No voom or count transformation is applied to continuous scores. The repeated-donor benchmark observed 84% interval coverage at 25 seeds (Monte Carlo SE 7.3%); calibration is not established.

```{r}
nonnegative <- fit_nonnegative_ebmf(abs(X),"rows",seed=12,max_factors=2,verbose=0,nullcheck=FALSE)
stopifnot(all(factor_activity(nonnegative)>=0),all(factor_effects(nonnegative)>=0))
```


## Source credits

Original function and package authors are named in each function manual and in
[Original authors and sources](credits.html). The installed `provenance()` registry
records source pins, authors, licenses and reuse. Please cite the upstream methods
used in your analysis; `citation("flashier")` and `citation("ebnm")` give core citations.
