---
title: "Compare groups and test associations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare groups and test associations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use group summaries to describe activity, or fit a regression to test an
association. In either case, metadata must identify samples by name.
The example uses simulated activities for 12 donors measured at two visits.

```{r setup, echo=FALSE}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48), 24, 2), matrix(rnorm(32), 16, 2)) +
  matrix(rnorm(384, sd = 0.2), 24, 16)
dimnames(X) <- list(paste0("s", 1:24), paste0("g", 1:16))
fit <- fit_ebmf(X, sample_side = "rows", seed = 1, max_factors = 2,
                backfit = TRUE, nullcheck = FALSE, verbose = 0,
                feature_scale = "simulated_intensity")
view <- standardize_factors(fit)
```

```{r metadata}
metadata <- data.frame(
  sample = rownames(X),
  donor = rep(paste0("d", 1:12), each = 2),
  group = factor(rep(c("A", "B"), each = 12)),
  visit = rep(0:1, 12)
)
```

## Describe group differences

```{r summaries}
summary <- summarize_factor_activity(view, metadata, by = "group", unit_col = "donor")
summary[, c("group", "factor", "n_units", "mean")]
difference <- compare_factor_groups(view, metadata, "group", c("B", "A"),
                                    unit_col = "donor")
difference$factor_contrast[, c("factor", "contrast", "estimate")]
```

The contrast is B minus A. These summaries have no p-values.
Means weight observations equally by default. `unit_col` counts donors but
does not balance their contributions when they have different numbers of observations.

## Test an association

For independent samples, use `lm`. Here, select one visit per donor and fit
that sample matrix.

```{r independent}
independent <- metadata[seq(1, 24, 2), ]
indfit <- fit_ebmf(X[independent$sample, ], sample_side = "rows", seed = 8,
                   max_factors = 2, verbose = 0, nullcheck = FALSE,
                   feature_scale = "simulated_intensity")
result <- test_factors(indfit, independent, ~group, engine = "lm",
                       unit_col = "donor", independence = "independent")
result$table[, c("factor", "term", "estimate", "std.error", "adj.p.value", "status")]
```

The table reports estimates, standard errors, p-values and adjusted p-values
for each factor. With A as the reference level, `groupB` estimates B minus A.
`engine = "limma"` provides moderated tests when limma is installed.
Moderation can be sensitive when there are few factors.

## Account for repeated donors

Install variancePartition to use `dream`. Include the declared donor in the
random effects and retain all visits.

```{r repeated}
if (requireNamespace("variancePartition", quietly = TRUE)) {
  repeated <- test_factors(
    view, metadata, ~visit + (1 | donor), engine = "dream",
    unit_col = "donor", independence = "repeated",
    control = list(dream = list(BPPARAM = BiocParallel::SerialParam()))
  )
  repeated$table[, c("factor", "term", "estimate", "std.error", "adj.p.value", "status")]
  repeated$diagnostics
}
```

Check `status` and `diagnostics` before interpreting a test.
All tests condition on the fitted activities and selected number of factors.
They do not propagate uncertainty from learning the factorization.
Treat donors as biological units when cells or visits share a donor.

To express a contrast in feature units, see
[Translate contrasts into feature effects](backprojection.html).

## Sources

Group summaries follow workflows by David Zemmour and the Zemmour Lab.
Tests call R Core Team's `lm`, Gordon Smyth and contributors' limma, or Gabriel
Hoffman's dream. [Original authors and sources](credits.html) lists citations.
