---
title: "Native transfer of activities and feature programs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Native transfer of activities and feature programs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(flashier.utils)
library(Matrix)
set.seed(9)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5)
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache_file <- file.path(pancreas_dir,"pancreas-worked-example.rds")
cells_file <- file.path(pancreas_dir,"pancreas-cells.rds")
have_pancreas <- nzchar(pancreas_dir) && file.exists(cache_file) && file.exists(cells_file)
knitr::opts_chunk$set(eval=have_pancreas)
```

```{r load-programs, include=FALSE}
# Restore published factors as in the pancreas-programs article (means only).
cache <- readRDS(cache_file)
source("restore-programs.R", local=knitr::knit_global())
view <- restore_programs(cache$reference)
```

Native projection delegates to flashier authors Jason Willwerscheid, Peter
Carbonetto, Wei Wang and Matthew Stephens. Reference-program transfer receives
conceptual credit to [STARCAT](https://github.com/immunogenomics/starCAT), Dylan
Kotliar and Michelle Curtis. This is independently written EBMF glue; it does not
introduce STARCAT query-specific scaling or normalized usages.

The examples use **published human-pancreas programs** from Peter Carbonetto's
[single-cell Jamboree](https://stephenslab.github.io/single-cell-jamboree/pancreas_annotate.html):
nine NMF programs learned from CEL-Seq2 cells (Muraro et al., 2016, as assembled
by Luecken et al., 2022), restored as `view` without refitting; see
[Pancreas programs](pancreas-programs.html). The reference measurements are 48
CEL-Seq2 cells (eight per published cell type, shifted-log expression of 14,528
genes). The code runs only when `FLASHIER_UTILS_PANCREAS_DIR` holds the local
caches written by `.github/scripts/prepare-pancreas.R` in the source repository
(`pancreas-worked-example.rds` and `pancreas-cells.rds`) and otherwise shows without
output. Upstream data are not redistributed.

## New measurements on the same cells

Here a small marker panel stands in for newly measured features: shifted-log
expression of eight genes in the same 48 cells. The genes are also among the
published programs, so the projection can be compared with the published
effects of the same genes.

```{r new-effects}
panel <- c("INS","IAPP","GCG","SST","PPY","REG1A","CTRB2","CFTR")
new_features <- as.matrix(cache$reference_expression[,panel])
projected <- project_factor_features(view,new_features,"rows",
  feature_scale="shifted_log_reference_library_normalized_measurements")
stopifnot(identical(projected$representation$activity,view$activity))
projected$reconstruction
# Estimated effect of each panel gene on its strongest published program.
strongest <- apply(view$effects[panel,],1,which.max)
data.frame(gene=panel,program=colnames(view$effects)[strongest],
  published=round(view$effects[cbind(panel,colnames(view$effects)[strongest])],2),
  projected=round(projected$representation$effects[cbind(panel,
    colnames(view$effects)[strongest])],2))
```

Sample IDs align measured cells; a subset of reference cells is permitted if
selected activities retain full rank. QR least squares initializes feature
weights, followed by native EBMF with activities fixed, feature-specific residual
variance, signed point-Laplace feature priors and 200 iterations by default.
A rank-deficient initialization fails with a factor-selection error. Projected
values are on the scale of the fixed activities. Here they track the published
coordinates for `INS` (1.02 against 1.00) and `IAPP`, but differ for others, such
as `SST` (0.57 against 1.00), because they are re-estimated from only 48 cells
and eight genes. Native messages, such as a factor estimate that is numerically
zero, are passed through.

## New cells with fixed reference programs

Query cells come from a different technology: inDrop (`inDrop3`) cells from the
same benchmark, eight per published cell type, with the CEL-Seq2 library-size mean
and the same 14,528 genes. Counts are transformed with the reference rule,
`log1p(count / library size * reference mean)`.

```{r new-cells}
cells <- readRDS(cells_file)
# Rows of the cached counts follow the order of the cached metadata.
stopifnot(nrow(cells$counts)==nrow(cells$metadata))
rownames(cells$counts) <- as.character(cells$metadata$id)
meta <- cells$metadata[cells$metadata$tech=="inDrop3",]
pick <- unlist(lapply(cells$cell_types,function(type)
  head(sort(as.character(meta$id[meta$celltype==type])),8)))
counts <- cells$counts[pick,rownames(view$effects),drop=FALSE]
totals <- Matrix::rowSums(counts)
stopifnot(all(totals>0))
query <- log1p(Matrix::Diagonal(x=cells$reference_library_mean/totals) %*% counts)
dimnames(query) <- dimnames(counts)
preprocessing <- list(scope="external_reference",revision=cache$revision,
  transformation="log1p(filtered measurement / library size * reference mean)",
  reference_library_mean=cache$reference_library_mean)
transfer <- project_factor_activity(view,query,"rows",
  feature_scale="shifted_log_reference_library_normalized_measurements",
  preprocessing=preprocessing)
stopifnot(identical(transfer$representation$effects,view$effects),
          identical(transfer$representation$offset,view$offset))
transfer$reconstruction
transfer$analysis_metadata$coverage
```

Declare compatible feature units and a preprocessing manifest. The caller is
responsible for applying the same reference transform; projection does not learn
or apply a transform. An equal transform does not establish biological
equivalence across technologies. As a descriptive check, mean scaled activity of
the transferred cells by their published cell type (five programs shown):

```{r transferred-activity}
query_info <- cells$metadata[match(rownames(transfer$representation$activity),
                                   cells$metadata$id),]
transferred <- sweep(transfer$representation$activity,2,
                     apply(view$activity,2,max),"/")
round(apply(transferred[,c("F2","F4","F5","F6","F9")],2,
            function(a) tapply(a,droplevels(query_info$celltype),mean)),2)
```

In this table the largest mean activity falls in `F2` for acinar, `F5` for alpha,
`F4` for beta, `F6` for delta and `F9` for ductal cells, matching the reference
programs, but the values are modest and gamma cells have no clear dedicated program.
The in-sample fit of the query (`fit_r2` above) is only moderate, and these
descriptions are not a validated cell-type assignment.

Matching defaults to exact gene sets. Explicit intersection records missing and
excluded genes, requires at least 50% reference coverage by default, and rejects
shared programs without full column rank. Here the first 2,500 published genes
are dropped to mimic a smaller measured panel.

```{r intersection}
intersection <- project_factor_activity(view,query[,-seq_len(2500)],"rows",
  feature_scale="shifted_log_reference_library_normalized_measurements",
  preprocessing=preprocessing, feature_match="intersection", min_coverage=0.8)
length(intersection$analysis_metadata$missing_reference_genes)
# One-to-one mapping names query genes and values identify reference genes.
mapped_query <- query
colnames(mapped_query) <- paste0("query_gene",seq_len(ncol(query)))
mapping <- setNames(colnames(query),colnames(mapped_query))
mapped <- project_factor_activity(view,mapped_query,"rows",
  feature_scale="shifted_log_reference_library_normalized_measurements",
  preprocessing=preprocessing, gene_mapping=mapping)
all.equal(mapped$representation$activity,transfer$representation$activity)
```

Source factor IDs, signs, order and fixed values are retained. Native priors are
inferred from recorded source metadata; bare/custom sources require explicit
`ebnm_fn` functions in native row/column order. For column-oriented samples use
`sample_side="columns"` and transpose the query. Noise and prior hyperparameters
may recalibrate on query observations, matching the holdout contract. Native
warnings and iteration-limit diagnostics remain in the output.

Sparse queries are supported where the native engine permits. Sparse plus
explicit `S` is refused. Nonzero offsets require dense centering; QR and scoring
allocations obey the default 256 MiB budget. Reconstruction metrics score all
shared query measurements in blocks; they are conditional fit diagnostics, not
independent validation. The returned representation retains even unmeasured
reference genes, without imputing them into the fitting data.

```{r serialization}
path <- tempfile(fileext=".rds")
saveRDS(view,path)
stopifnot(identical(readRDS(path),view))
unlink(path)
```

Use ordinary R objects and `saveRDS()` to retain the reference and preprocessing
manifest. See [holdout evaluation](holdout.html) for independent prediction and
[credits](credits.html) for upstream authors and citations.

## Published-data example

See [Pancreas programs from the single-cell Jamboree](pancreas-programs.html)
for the factor 4-6 comparison, diagnostics and transfer to SMART-Seq2 cells. This
article uses the same CEL-Seq2 programs (credit: Peter Carbonetto's single-cell
Jamboree; Muraro et al., 2016; Luecken et al., 2022) and the benchmark's inDrop
cells to teach the transfer functions. The article runs when the local data
archive is configured, as in the website build; package checks skip it.
