---
title: "Reorder, remove, fix and refit factors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reorder, remove, fix and refit factors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

An EBMF fit is a set of factors that you can edit. This tutorial reorders and
removes factors, holds one factor fixed while the others are refined, and checks
how much of the data each factor explains. Each function returns a new fit and
leaves the original unchanged.

## Start from a fit

Use the GTEx example from flashier, loaded with `data("gtex", package = "flashier")` ([documentation](https://willwerscheid.github.io/flashier/reference/gtex.html)). Rows pair a single nucleotide polymorphism
(SNP) with a gene, columns are tissues, and entries are association z-scores.

```{r fit}
library(flashier.utils)
data("gtex", package = "flashier")
fit <- fit_ebmf(gtex, sample_side = "columns", max_factors = 3,
                backfit = TRUE, seed = 1, verbose = 0,
                feature_scale = "association_z_score")
colnames(factor_activity(fit))
```

## Reorder factors

`reorder_factors()` takes an exact permutation of the retained factor IDs.

```{r reorder}
reordered <- reorder_factors(fit, c("F3", "F1", "F2"))
colnames(factor_activity(reordered))
```

The IDs travel with their activity and feature effects. `F3` is still the factor
that is strongest in whole blood; only its position changes.

## Remove factors

`remove_factors()` drops factors by ID.

```{r remove}
smaller <- remove_factors(fit, "F1")
colnames(factor_activity(smaller))
```

The remaining factors keep their original IDs, here `F2` and `F3`. IDs are not
renumbered, so results saved before the removal still refer to the same factors.
Removing a factor changes the reconstruction, which no longer includes it. Refit
afterwards if you want the remaining factors to absorb what it explained.

## Fix a factor, then refit the others

`fix_factors()` holds one side of a factor constant during later refinement. Fix
either the feature effects or the activity of a factor; flashier 1.0.60 errors
if both are fixed for the same factor.

```{r fix-refit}
fixed <- fix_factors(fit, "F2", side = "effects")
refit <- refit_factors(fixed, factors = c("F1", "F3"),
                       control = list(maxiter = 20, verbose = 0), seed = 1)
max(abs(factor_effects(refit, factors = "F2") -
        factor_effects(fit, factors = "F2")))
```

The difference is zero: the fixed effects were not changed. `refit_factors()`
backfits only the listed factors, using a scoped seed. Use
`fix_factors(..., action = "unfix")` to release a factor. Fixing is how you hold
a published or previously fitted factor constant while estimating new ones.

## Check variance explained

`factor_pve()` reports flashier's proportion of variance explained by each factor.

```{r pve}
factor_pve(fit)
factor_pve(refit)
```

The first factor explains most of the variance because it captures effects that
are shared across tissues. A factor that explains little can still be
biologically specific, as `F2` is for brain tissues. Proportions of variance
describe the fitted matrix, not the evidence for a factor. Refitting the other
two factors leaves the proportions almost unchanged here.

## Sources

The editing functions wrap `flash_factors_reorder`, `flash_factors_remove`,
`flash_factors_fix`, `flash_factors_unfix` and `flash_backfit`, and `factor_pve`
reads the native proportion of variance explained, all from flashier by
Jason Willwerscheid, Peter Carbonetto, Wei Wang and Matthew Stephens. The GTEx
data ship with flashier and derive from the GTEx project (see `?flashier::gtex`).
[Original authors and sources](credits.html) lists the full credits and citations.

## Function credits

These are the functions from this package that the article calls, with the people
who wrote the upstream code or method behind each one. Use is this package's
classification and code reuse for the function; `provenance()` returns the full
registry. [Original authors and sources](credits.html) gives the full credits,
citations and licenses.

```{r function-credits, echo=FALSE, eval=TRUE}
code <- knitr::knit_code$get()
used <- unique(unlist(lapply(code[names(code) != "function-credits"], function(x)
  tryCatch(all.names(parse(text = x)), error = function(e) NULL))))
registry <- provenance()
credits <- registry[registry[["function"]] %in% used, ]
credits$use <- paste(credits$classification, credits$code_reuse, sep = " / ")
knitr::kable(credits[, c("function", "authors", "conceptual_origin", "use")],
             col.names = c("Function", "Original authors", "Upstream", "Use"),
             row.names = FALSE)
```
