---
title: "Annotate factors with pathways"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Annotate factors with pathways}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Annotate a factor from its feature effects using gene sets or signed networks.
Install fgsea for ranked enrichment, or decoupleR for network scores.
The examples use simulated feature IDs and annotations with no biological meaning.

```{r setup, echo=FALSE}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48), 24, 2), matrix(rnorm(32), 16, 2)) +
  matrix(rnorm(384, sd = 0.2), 24, 16)
dimnames(X) <- list(paste0("s", 1:24), paste0("g", 1:16))
fit <- fit_ebmf(X, sample_side = "rows", seed = 1, max_factors = 2,
                backfit = TRUE, nullcheck = FALSE, verbose = 0,
                feature_scale = "simulated_intensity")
view <- standardize_factors(fit)
```

## Gene-set enrichment

Supply a named list of feature sets. Set members must match the fit's feature IDs.

```{r gene-sets}
pathways <- list(panelA = paste0("g", 1:6), panelB = paste0("g", 7:12))
if (requireNamespace("fgsea", quietly = TRUE)) {
  enrichment <- enrich_factors(
    view, pathways, engine = "fgsea", factors = "F1",
    control = list(seed = 12, nPermSimple = 100L, eps = 1e-3, minSize = 3L)
  )
  enrichment$table[, c("pathway", "factor", "NES", "padj")]
}
```

fgsea ranks features by their signed effects and reports enrichment statistics,
p-values and adjusted p-values. For an analysis, record the gene-set source,
organism and version. The small permutation budget here keeps the example short.

## Signed network scores

A network needs `source` (program), `target` (feature ID) and `mor` (signed weight).
The univariate linear model (ULM) scores each program separately.
The multivariate linear model (MLM) fits programs jointly.

```{r network}
network <- data.frame(
  source = rep(c("programA", "programB"), each = 6),
  target = paste0("g", 1:12),
  mor = rep(c(1, -1, 0.5, -0.5, 2, -2), 2)
)
attr(network, "annotation_version") <- "simulated_v1"
if (requireNamespace("decoupleR", quietly = TRUE)) {
  annotation <- enrich_factors(view, network, engine = "decoupleR",
                               control = list(method = "ulm", minsize = 3L))
  annotation$table[, c("source", "condition", "score", "p_value")]
  annotation$analysis_metadata$pathway_sizes
}
```

`condition` identifies the factor. Use `method = "mlm"` for joint scoring.
Gene networks such as CollecTRI or PROGENy require gene-level features.
For PROGENy, copy its `weight` column to `mor`. For CyTOF, supply a network
that describes the measured markers.

## Read the results

Check feature overlap in `analysis_metadata$pathway_sizes`.
Use `feature_map` for an explicit one-to-one ID mapping if needed.
Duplicate source-target edges are rejected.

Network scores annotate feature programs. They do not measure pathway activity
in individual samples. Scores depend on factor orientation, and native p-values
do not account for uncertainty in the factorization.
The default leaves feature effects uncentered across factors.

## Sources

The gene-set workflow credits Peter Carbonetto and Matthew Stephens.

fgsea is by Gennady Korotkevich, Vladimir Sukhov and Alexey Sergushichev.

decoupleR is by Pau Badia-i-Mompel, Jesús Vélez-Santiago and colleagues.
[Original authors and sources](credits.html) lists all authors and method citations.
