---
title: "Annotate factors with pathways"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Annotate factors with pathways}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(flashier.utils)
library(Matrix)
set.seed(3)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5)
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache_file <- file.path(pancreas_dir,"pancreas-worked-example.rds")
cells_file <- file.path(pancreas_dir,"pancreas-cells.rds")
have_pancreas <- nzchar(pancreas_dir) && file.exists(cache_file) && file.exists(cells_file)
knitr::opts_chunk$set(eval=have_pancreas)
```

Annotate a factor from its feature effects using gene sets or signed networks.
Install fgsea for ranked enrichment, or decoupleR for network scores.

## Data and objects

**Data.** Nine non-negative matrix factorization (NMF) programs learned from CEL-Seq2 cells (Muraro et al., 2016, as assembled by Luecken et al., 2022), published in [Peter Carbonetto's single-cell Jamboree](https://stephenslab.github.io/single-cell-jamboree/pancreas_annotate.html). The article uses NMF programs (`view`) and semi-NMF programs with signed effects (`signed_view`), both restored without refitting. Their effects cover 14,528 genes. The gene sets are built from cells of three other technologies in the same benchmark, as shown below.

**Get the data.** The package does not ship these data. From a clone of the
[repository](https://github.com/mdmanurung/flashier-utils), run this once; it downloads about 86 MB from a pinned
revision of the Jamboree repository ([details](pancreas-programs.html#prepare-a-local-example-cache)):

```sh
FLASHIER_UTILS_PANCREAS_DIR=~/pancreas-jamboree-example Rscript .github/scripts/prepare-pancreas.R
```

This article reads `pancreas-worked-example.rds` and, for the gene sets, `pancreas-cells.rds`. Set `FLASHIER_UTILS_PANCREAS_DIR` to the same directory before running the
code below; without it the code chunks are skipped. This website was built with the data.

**Objects.** `cache` is the list read from `pancreas-worked-example.rds`: the published programs (`reference`, and `signed` for the semi-NMF programs), 48 reference cells (eight per published cell type), their shifted-log expression and metadata. `view` and `signed_view` are the NMF and semi-NMF representations. `cells` is read from `pancreas-cells.rds` where it is first used. `restore_programs()` is defined in [restore-programs.R](https://github.com/mdmanurung/flashier-utils/blob/main/vignettes/restore-programs.R) and explained in [Pancreas programs](pancreas-programs.html#restore-published-coordinates-without-refitting).

```{r load-programs}
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache <- readRDS(file.path(pancreas_dir, "pancreas-worked-example.rds"))
source("restore-programs.R")  # from vignettes/ in the repository
view <- restore_programs(cache$reference, cache)
signed_view <- restore_programs(cache$signed, cache, signed_effects = TRUE)
```

No external pathway database is bundled or used here. Instead, the gene sets are
**marker sets built from the published cell-type labels**, in cells from three other
technologies of the same benchmark (SMART-Seq2, Fluidigm C1 and inDrop), so they are
defined without the CEL-Seq2 cells that the programs were learned from. They are not
pathways, and the result is a check of agreement between programs and
label-defined markers. These are different cells measured with different
technologies, but the data carry no donor identifiers, so no donor-level
independence is claimed.

```{r marker-sets}
cells <- readRDS(file.path(pancreas_dir, "pancreas-cells.rds"))
stopifnot(nrow(cells$counts)==nrow(cells$metadata))
rownames(cells$counts) <- as.character(cells$metadata$id)
others <- c("smartseq2","fluidigmc1","inDrop3")
meta <- cells$metadata[cells$metadata$tech %in% others,]
counts <- cells$counts[as.character(meta$id),rownames(view$effects)]
Z <- log1p(Matrix::Diagonal(x=cells$reference_library_mean/Matrix::rowSums(counts)) %*% counts)
# Mean shifted-log difference between a cell type and all other types,
# averaged over the three technologies.
type_contrast <- function(type) rowMeans(vapply(others,function(tech) {
  use <- meta$tech==tech
  Matrix::colMeans(Z[use & meta$celltype==type,,drop=FALSE]) -
    Matrix::colMeans(Z[use & meta$celltype!=type,,drop=FALSE])
},numeric(ncol(Z))))
types <- c("acinar","ductal","alpha","beta","delta","gamma")
contrast <- sapply(types,type_contrast)
pathways <- setNames(lapply(types,function(type)
  names(sort(contrast[,type],decreasing=TRUE))[1:50]),paste0(types,"_markers"))
vapply(pathways,function(set) paste(head(set,4),collapse=", "),"")
```

## Gene-set enrichment

Supply a named list of feature sets. Set members must match the fit's feature IDs.

```{r gene-sets}
have_fgsea <- requireNamespace("fgsea", quietly = TRUE)
if (have_fgsea)
  enrichment <- enrich_factors(
    view, pathways, engine = "fgsea", factors = c("F2","F4","F5","F9"),
    control = list(seed = 12, scoreType = "pos", nPermSimple = 1000L, eps = 0,
                   minSize = 10L)
  )
if (have_fgsea) enrichment$table[, c("pathway", "factor", "NES", "padj")]
```

fgsea ranks features by their signed effects and reports enrichment statistics,
p-values and adjusted p-values. For an analysis, record the gene-set source,
organism and version; here the source is the marker construction above.
The effects are non-negative, so `scoreType = "pos"` is used. For each program the
set with the largest normalized enrichment score (NES) is the one expected from the
program's top genes: acinar markers for `F2` (NES 1.45), beta markers for `F4`
(1.42), alpha markers for `F5` (1.32) and ductal markers for `F9` (1.68).
The margins are small, however. `F4` is also enriched for delta, gamma and alpha
markers (NES 1.35 to 1.40) and `F2` for ductal markers (1.31), as these sets share
endocrine or exocrine genes, so the table narrows candidates rather than assigning
a cell type. Read it together with each program's top genes (`factor_features()`).

## Signed network scores

A network needs `source` (program), `target` (feature ID) and `mor` (signed weight).
The univariate linear model (ULM) scores each program separately.
The multivariate linear model (MLM) fits programs jointly. Here each source is a
cell type, with the 50 genes highest and the 25 genes lowest in its contrast, signed
by the mean expression difference (`mor`).

```{r network}
network <- do.call(rbind,lapply(types,function(type) {
  up <- names(sort(contrast[,type],decreasing=TRUE))[1:50]
  down <- names(sort(contrast[,type]))[1:25]
  data.frame(source=type,target=c(up,down),mor=contrast[c(up,down),type])
}))
rownames(network) <- NULL
attr(network, "annotation_version") <- "marker contrasts v1 (smartseq2, fluidigmc1, inDrop3)"
have_decoupler <- requireNamespace("decoupleR", quietly = TRUE)
if (have_decoupler)
  annotation <- enrich_factors(signed_view, network, engine = "decoupleR",
                               factors = c("F2","F3","F4","F8"),
                               control = list(method = "ulm", minsize = 3L))
if (have_decoupler) annotation$table[, c("source", "condition", "score", "p_value")]
if (have_decoupler) annotation$analysis_metadata$pathway_sizes
```

`condition` identifies the factor. These are the semi-NMF programs of
`signed_view`, whose IDs need not match the NMF IDs used above. The highest score of
each shown program is for `F2` acinar (46.1) and ductal (44.8) with negative scores
for the endocrine sets, `F3` alpha (55.1), `F4` beta (40.8) and `F8` delta (37.0),
with gamma second for `F3` (24.0) and `F8` (16.6). The ULM p-values treat genes as
independent observations and are extremely small throughout, so use the score and
its sign rather than the p-value as the summary. Use `method = "mlm"` for joint scoring.
Gene networks such as CollecTRI or PROGENy require gene-level features.
For PROGENy, copy its `weight` column to `mor`. For CyTOF, supply a network
that describes the measured markers.

## Read the results

Check feature overlap in `analysis_metadata$pathway_sizes`.
Use `feature_map` for an explicit one-to-one ID mapping if needed.
Duplicate source-target edges are rejected.

Network scores annotate feature programs. They do not measure pathway activity
in individual samples. Scores depend on factor orientation, and native p-values
do not account for uncertainty in the factorization.
The default leaves feature effects uncentered across factors.

## Sources

The gene-set workflow credits Peter Carbonetto and Matthew Stephens.

fgsea is by Gennady Korotkevich, Vladimir Sukhov and Alexey Sergushichev.

decoupleR is by Pau Badia-i-Mompel, Jesús Vélez-Santiago and colleagues.
[Original authors and sources](credits.html) lists all authors and method citations.

The pancreas data and factors come from Peter Carbonetto's single-cell Jamboree
(CEL-Seq2 cells from Muraro et al., 2016, as assembled by Luecken et al., 2022);
native EBMF methods credit Jason Willwerscheid, Peter Carbonetto, Wei Wang and
Matthew Stephens. The marker sets were constructed here from the published
cell-type labels.

## Function credits

These are the functions from this package that the article calls, with the people
who wrote the upstream code or method behind each one. Use is this package's
classification and code reuse for the function; `provenance()` returns the full
registry. [Original authors and sources](credits.html) gives the full credits,
citations and licenses.

```{r function-credits, echo=FALSE, eval=TRUE}
code <- knitr::knit_code$get()
used <- unique(unlist(lapply(code[names(code) != "function-credits"], function(x)
  tryCatch(all.names(parse(text = x)), error = function(e) NULL))))
registry <- provenance()
credits <- registry[registry[["function"]] %in% used, ]
credits$use <- paste(credits$classification, credits$code_reuse, sep = " / ")
knitr::kable(credits[, c("function", "authors", "conceptual_origin", "use")],
             col.names = c("Function", "Original authors", "Upstream", "Use"),
             row.names = FALSE)
```
