---
title: "Annotate factors with pathways"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Annotate factors with pathways}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(flashier.utils)
library(Matrix)
set.seed(3)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5)
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache_file <- file.path(pancreas_dir,"pancreas-worked-example.rds")
cells_file <- file.path(pancreas_dir,"pancreas-cells.rds")
have_pancreas <- nzchar(pancreas_dir) && file.exists(cache_file) && file.exists(cells_file)
knitr::opts_chunk$set(eval=have_pancreas)
```

```{r load-programs, include=FALSE}
# Restore published factors as in the pancreas-programs article (means only).
cache <- readRDS(cache_file)
source("restore-programs.R", local=knitr::knit_global())
view <- restore_programs(cache$reference)
signed_view <- restore_programs(cache$signed,signed_effects=TRUE)
```

Annotate a factor from its feature effects using gene sets or signed networks.
Install fgsea for ranked enrichment, or decoupleR for network scores.

The examples use the **published human-pancreas programs** from Peter Carbonetto's
[single-cell Jamboree](https://stephenslab.github.io/single-cell-jamboree/pancreas_annotate.html):
NMF programs (`view`) and semi-NMF programs with signed effects (`signed_view`)
learned from CEL-Seq2 cells (Muraro et al., 2016, as assembled by Luecken et al.,
2022), restored without refitting as described in
[Pancreas programs](pancreas-programs.html). Their effects cover 14,528 genes. The
code runs only when `FLASHIER_UTILS_PANCREAS_DIR` holds the local caches written by
`.github/scripts/prepare-pancreas.R` in the source repository and otherwise shows
without output. Upstream data are not redistributed.

No external pathway database is bundled or used here. Instead, the gene sets are
**marker sets built from the published cell-type labels**, in cells from three other
technologies of the same benchmark (SMART-Seq2, Fluidigm C1 and inDrop), so they are
defined without the CEL-Seq2 cells that the programs were learned from. They are not
pathways, and the result is a check of agreement between programs and
label-defined markers. These are different cells measured with different
technologies, but the data carry no donor identifiers, so no donor-level
independence is claimed.

```{r marker-sets}
cells <- readRDS(cells_file)
stopifnot(nrow(cells$counts)==nrow(cells$metadata))
rownames(cells$counts) <- as.character(cells$metadata$id)
others <- c("smartseq2","fluidigmc1","inDrop3")
meta <- cells$metadata[cells$metadata$tech %in% others,]
counts <- cells$counts[as.character(meta$id),rownames(view$effects)]
Z <- log1p(Matrix::Diagonal(x=cells$reference_library_mean/Matrix::rowSums(counts)) %*% counts)
# Mean shifted-log difference between a cell type and all other types,
# averaged over the three technologies.
type_contrast <- function(type) rowMeans(vapply(others,function(tech) {
  use <- meta$tech==tech
  Matrix::colMeans(Z[use & meta$celltype==type,,drop=FALSE]) -
    Matrix::colMeans(Z[use & meta$celltype!=type,,drop=FALSE])
},numeric(ncol(Z))))
types <- c("acinar","ductal","alpha","beta","delta","gamma")
contrast <- sapply(types,type_contrast)
pathways <- setNames(lapply(types,function(type)
  names(sort(contrast[,type],decreasing=TRUE))[1:50]),paste0(types,"_markers"))
vapply(pathways,function(set) paste(head(set,4),collapse=", "),"")
```

## Gene-set enrichment

Supply a named list of feature sets. Set members must match the fit's feature IDs.

```{r gene-sets}
have_fgsea <- requireNamespace("fgsea", quietly = TRUE)
if (have_fgsea)
  enrichment <- enrich_factors(
    view, pathways, engine = "fgsea", factors = c("F2","F4","F5","F9"),
    control = list(seed = 12, scoreType = "pos", nPermSimple = 1000L, eps = 0,
                   minSize = 10L)
  )
if (have_fgsea) enrichment$table[, c("pathway", "factor", "NES", "padj")]
```

fgsea ranks features by their signed effects and reports enrichment statistics,
p-values and adjusted p-values. For an analysis, record the gene-set source,
organism and version; here the source is the marker construction above.
The effects are non-negative, so `scoreType = "pos"` is used. For each program the
set with the largest normalized enrichment score (NES) is the one expected from the
program's top genes: acinar markers for `F2` (NES 1.45), beta markers for `F4`
(1.42), alpha markers for `F5` (1.32) and ductal markers for `F9` (1.68).
The margins are small, however. `F4` is also enriched for delta, gamma and alpha
markers (NES 1.35 to 1.40) and `F2` for ductal markers (1.31), as these sets share
endocrine or exocrine genes, so the table narrows candidates rather than assigning
a cell type. Read it together with each program's top genes (`factor_features()`).

## Signed network scores

A network needs `source` (program), `target` (feature ID) and `mor` (signed weight).
The univariate linear model (ULM) scores each program separately.
The multivariate linear model (MLM) fits programs jointly. Here each source is a
cell type, with the 50 genes highest and the 25 genes lowest in its contrast, signed
by the mean expression difference (`mor`).

```{r network}
network <- do.call(rbind,lapply(types,function(type) {
  up <- names(sort(contrast[,type],decreasing=TRUE))[1:50]
  down <- names(sort(contrast[,type]))[1:25]
  data.frame(source=type,target=c(up,down),mor=contrast[c(up,down),type])
}))
rownames(network) <- NULL
attr(network, "annotation_version") <- "marker contrasts v1 (smartseq2, fluidigmc1, inDrop3)"
have_decoupler <- requireNamespace("decoupleR", quietly = TRUE)
if (have_decoupler)
  annotation <- enrich_factors(signed_view, network, engine = "decoupleR",
                               factors = c("F2","F3","F4","F8"),
                               control = list(method = "ulm", minsize = 3L))
if (have_decoupler) annotation$table[, c("source", "condition", "score", "p_value")]
if (have_decoupler) annotation$analysis_metadata$pathway_sizes
```

`condition` identifies the factor. These are the semi-NMF programs of
`signed_view`, whose IDs need not match the NMF IDs used above. The highest score of
each shown program is for `F2` acinar (46.1) and ductal (44.8) with negative scores
for the endocrine sets, `F3` alpha (55.1), `F4` beta (40.8) and `F8` delta (37.0),
with gamma second for `F3` (24.0) and `F8` (16.6). The ULM p-values treat genes as
independent observations and are extremely small throughout, so use the score and
its sign rather than the p-value as the summary. Use `method = "mlm"` for joint scoring.
Gene networks such as CollecTRI or PROGENy require gene-level features.
For PROGENy, copy its `weight` column to `mor`. For CyTOF, supply a network
that describes the measured markers.

## Read the results

Check feature overlap in `analysis_metadata$pathway_sizes`.
Use `feature_map` for an explicit one-to-one ID mapping if needed.
Duplicate source-target edges are rejected.

Network scores annotate feature programs. They do not measure pathway activity
in individual samples. Scores depend on factor orientation, and native p-values
do not account for uncertainty in the factorization.
The default leaves feature effects uncentered across factors.

## Sources

The gene-set workflow credits Peter Carbonetto and Matthew Stephens.

fgsea is by Gennady Korotkevich, Vladimir Sukhov and Alexey Sergushichev.

decoupleR is by Pau Badia-i-Mompel, Jesús Vélez-Santiago and colleagues.
[Original authors and sources](credits.html) lists all authors and method citations.

The pancreas data and factors come from Peter Carbonetto's single-cell Jamboree
(CEL-Seq2 cells from Muraro et al., 2016, as assembled by Luecken et al., 2022);
native EBMF methods credit Jason Willwerscheid, Peter Carbonetto, Wei Wang and
Matthew Stephens. The marker sets were constructed here from the published
cell-type labels.
