---
title: "Fit and interpret factors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fit and interpret factors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A factor links a pattern across samples to a pattern across features.
This tutorial fits a matrix, plots sample activity and finds features that
contribute to a factor.

## Fit a matrix

Use the GTEx example from flashier. Its rows pair a single nucleotide polymorphism (SNP) with a gene. Its columns
are tissues, and its entries are association z-scores.

```{r fit}
library(flashier.utils)
data("gtex", package = "flashier")

fit <- fit_ebmf(gtex, sample_side = "columns", max_factors = 3,
                backfit = TRUE, seed = 1, verbose = 0,
                feature_scale = "association_z_score")
```

`max_factors` sets an upper limit. flashier can retain fewer factors.
`backfit = TRUE` jointly refines the factors after adding them.

For your own data, use a numeric matrix with unique row and column names.
Choose `sample_side = "rows"` if samples occupy rows.
Preprocess the matrix before fitting and record its units in `feature_scale`.

## Plot sample activity

`standardize_factors()` returns matching activity and feature-effect matrices.
Activity always has samples in rows, even when the input has samples in columns.

```{r activity, fig.width=7, fig.height=8}
view <- standardize_factors(fit)
plot_factor_activity(view)
```

Each column is a factor and each row is a tissue. Colors show signed activity.
F2 separates brain tissues from many other tissues.
F3 has its strongest activity in whole blood.
Reversing both a factor's activity and its feature effects leaves the fitted
matrix unchanged. The sign alone does not identify activation or repression.

## Find contributing features

```{r features}
features <- factor_features(view, factor = "F1", select = "largest", n = 5)
features[, c("feature", "estimate")]
```

`estimate` is the signed feature effect. The largest absolute effects identify
the strongest contributors. Here, the feature IDs refer to SNP-gene pairs.
For a gene-expression matrix, they would be the gene IDs you supplied.

## Fit non-negative factors

When activity and feature effects should both be non-negative, use
`fit_nonnegative_ebmf()`. Association z-scores are signed, so fit their absolute
values, which measure the strength of each association and drop its direction.

```{r nonnegative}
strength <- fit_nonnegative_ebmf(abs(gtex), sample_side = "columns",
                                 max_factors = 3, seed = 1, verbose = 0,
                                 feature_scale = "absolute_association_z_score")
round(factor_activity(strength)[c("Brain_Cortex", "Testis", "Whole_Blood"), ], 2)
```

`F1` is a shared strength pattern, `F3` is active only in brain tissues and `F2`
only in testis. Activity is never negative, so a factor adds strength and cannot
cancel it. The non-negative fit uses a native greedy start and a fixed prior
recipe; see [Reorder, remove, fix and refit factors](edit-fits.html) to edit it.

See [Find defining features](largest-distinctive.html) for rankings and plots.
Use [Compare groups and test associations](contrasts.html) to relate factors to sample metadata.
These tissue summaries do not provide donor-level replication.

## Sources

This example follows the [flashier introduction](https://willwerscheid.github.io/flashier/articles/flashier_intro.html).
flashier is by Jason Willwerscheid, Peter Carbonetto, Wei Wang and Matthew Stephens.
The GTEx data come from Lonsdale et al. (2013), with the selection described by
Urbut, Wang, Carbonetto and Stephens (2019). See the
[GTEx data reference](https://willwerscheid.github.io/flashier/reference/gtex.html).

Feature ranking adapts singlecelljamboreeR helpers by Peter Carbonetto and
Matthew Stephens. The display workflow credits David Zemmour.
[Original authors and sources](credits.html) lists the full credits and citations.
