---
title: "First signed fit to interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{First signed fit to interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(24),12,2),matrix(rnorm(16),8,2)) + matrix(rnorm(96,sd=0.2),12,8)
dimnames(X) <- list(paste0("s",1:12),paste0("g",1:8))
fit <- fit_ebmf(X,"rows",seed=1,max_factors=2,verbose=0,feature_scale="arcsinh_intensity")
view <- standardize_factors(fit)
factor_features(view,"F1",n=3)
plot_factor_activity(view)
```

Rows are samples, not implicitly independent donors. No biological replication
or inferential p-value is established by this example. The same interface takes
a named, user-preprocessed pseudobulk matrix with its scale and independent unit
recorded. Signed scores stay signed.

```{r}
beta <- setNames(rep(1,ncol(view$activity)),colnames(view$activity))
backproject_contrast(view,beta,"coefficients",factor_basis=view$manifest)$effects
factor_uncertainty(fit, "activity")
provenance("backproject_contrast")
```

These feature effects represent a declared activity contrast in arcsinh units.
Factor posterior SD is conditional on fitted EBMF priors and differs from
regression standard error. This workflow credits flashier, singlecelljamboreeR
and the ZemmourLib/ImmGenT backprojection recipe; see the installed NOTICE.


## Source credits

Original function and package authors are named in each function manual and in
[Original authors and sources](credits.html). The installed `provenance()` registry
records source pins, authors, licenses and reuse. Please cite the upstream methods
used in your analysis; `citation("flashier")` and `citation("ebnm")` give core citations.
