---
title: "Evaluate held-out prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluate held-out prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Learn feature programs on training samples, then evaluate held-out donors.
Use projection to assess reconstruction from observed entries, or masked
prediction to score features excluded from activity estimation.
The example uses simulated data with two visits per donor.

```{r setup, echo=FALSE}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48), 24, 2), matrix(rnorm(32), 16, 2)) +
  matrix(rnorm(384, sd = 0.2), 24, 16)
dimnames(X) <- list(paste0("s", 1:24), paste0("g", 1:16))
fit <- fit_ebmf(X, sample_side = "rows", seed = 1, max_factors = 2,
                backfit = TRUE, nullcheck = FALSE, verbose = 0,
                feature_scale = "simulated_intensity")
view <- standardize_factors(fit)
```

## Define a split

Keep all observations from a donor in the same fold.

```{r folds}
metadata <- data.frame(sample = rownames(X),
                       donor = rep(paste0("d", 1:12), each = 2))
folds <- data.frame(sample = metadata$sample,
                    fold = rep(c("training", "held"), c(16, 8)))
fit_training <- function(X, metadata, sample_side, seed, settings,
                         observation_aux = list()) {
  fit_ebmf(X, sample_side = sample_side, seed = seed, max_factors = 2,
            verbose = 0, backfit = TRUE, nullcheck = FALSE,
            feature_scale = "simulated_intensity")
}
control <- list(sample_side = "rows",
                preprocessing_manifest = list(scope = "identity"),
                evaluated_folds = "held")
```

`identity` declares that this example performs no preprocessing.
For an analysis, estimate normalization on training data and apply it to test
data. If preprocessing uses all samples, label the results as conditional on
that global preprocessing.

## Project held-out samples

```{r projection}
projection <- factor_holdout(X, metadata, "donor", folds, fit_training,
                             evaluation = "projection", seed = 1, control = control)
projection$metrics
```

Feature programs stay fixed. The method estimates test activities from observed
test entries. Noise and prior parameters can recalibrate on those entries.
This evaluates reconstruction after observing the test features.

## Predict masked features

```{r masked}
masked <- factor_holdout(
  X, metadata, "donor", folds, fit_training, evaluation = "masked",
  activity_features = paste0("g", 1:8), scoring_features = paste0("g", 9:16),
  seed = 1, control = control
)
masked$metrics
```

The two feature sets must be disjoint. Scoring features are excluded from test
activity estimation and noise or prior calibration. This is the appropriate
comparison when the question concerns prediction of unobserved features.

For a cohort split, use cohort labels as folds and donor IDs as units.
Each donor must belong to one cohort.

## Sources

Projection uses public flashier functions by Jason Willwerscheid,
Peter Carbonetto, Wei Wang and Matthew Stephens.
[Original authors and sources](credits.html) lists method citations.
