---
title: "Find defining features"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Find defining features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use feature effects to identify measurements that describe a factor.
The examples use a simulated fit, `view`, with two factors and 16 features.

```{r setup, echo=FALSE}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48), 24, 2), matrix(rnorm(32), 16, 2)) +
  matrix(rnorm(384, sd = 0.2), 24, 16)
dimnames(X) <- list(paste0("s", 1:24), paste0("g", 1:16))
fit <- fit_ebmf(X, sample_side = "rows", seed = 1, max_factors = 2,
                backfit = TRUE, nullcheck = FALSE, verbose = 0,
                feature_scale = "simulated_intensity")
view <- standardize_factors(fit)
```

## Rank features

```{r rankings}
largest <- factor_features(view, factor = "F1", select = "largest", n = 4)
largest[, c("feature", "estimate")]

distinctive <- factor_features(view, factor = "F1", select = "distinctive", n = 4)
distinctive[, c("feature", "estimate", "distinctiveness")]
```

`largest` selects the strongest absolute effects.
`distinctive` selects effects that exceed the most extreme same-direction
effect among competing factors, with zero included as a competitor.
A feature can contribute strongly to several factors without distinguishing one.
The table keeps its signed effect in `estimate` and its competitor margin in
`distinctiveness`.

## Plot selected features

```{r plot, fig.width=6, fig.height=4}
plot_factor_features(view, factors = "F1", select = "largest", n = 4)
```

Colors show the signed feature effects. Rankings are descriptive and depend
on the factor scale and sign convention. Keep the same representation when
comparing factors. See [Factor scales and measurement units](representation-units.html).

## Sources

The ranking functions adapt `rank_effects` and `compute_le_effects` from
singlecelljamboreeR by Peter Carbonetto and Matthew Stephens.
The plot follows a workflow by David Zemmour.
[Original authors and sources](credits.html) lists licenses and citations.
