---
title: "Find defining features"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Find defining features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(flashier.utils)
library(Matrix)
set.seed(3)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=6, fig.height=4)
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache_file <- file.path(pancreas_dir,"pancreas-worked-example.rds")
have_pancreas <- nzchar(pancreas_dir) && file.exists(cache_file)
knitr::opts_chunk$set(eval=have_pancreas)
```

```{r load-programs, include=FALSE}
# Restore published factors as in the pancreas-programs article (means only).
cache <- readRDS(cache_file)
ldf <- cache$reference
Y <- cache$reference_expression
factor_ids <- paste0("F",seq_len(ncol(ldf$F)))
A <- sweep(ldf$L[rownames(Y),,drop=FALSE],2,ldf$D,"*")
B <- ldf$F
colnames(A) <- colnames(B) <- factor_ids
priors <- list(ebnm::ebnm_point_exponential,ebnm::ebnm_point_exponential)
object <- fit_ebmf(Y,"rows",ebnm_fn=priors,max_factors=0,var_type=2,seed=1,
  feature_scale="shifted_log_reference_library_normalized_measurements",verbose=0)
metadata <- attr(object,"flashbridge_metadata")
metadata$factor_ids <- factor_ids
metadata$preprocessing <- list(scope="external_reference",revision=cache$revision,
  transformation="log1p(filtered measurement / library size * reference mean)",
  reference_library_mean=cache$reference_library_mean,
  source_sha256=cache$source_sha256)
object <- flashier::flash_factors_init(object,list(A,B),ebnm_fn=priors)
object <- flashier::flash_factors_fix(object,seq_along(factor_ids),"factors")
attr(object,"flashbridge_metadata") <- metadata
view <- standardize_factors(object,sample_side="rows",scaling="raw",
  prior_support=c(activity="nonnegative",effects="nonnegative"))
stopifnot(identical(unname(view$effects),unname(B)))
```

Use feature effects to identify measurements that describe a factor. The
examples use **published human-pancreas programs** from Peter Carbonetto's
[single-cell Jamboree](https://stephenslab.github.io/single-cell-jamboree/pancreas_annotate.html):
nine non-negative matrix factors learned from CEL-Seq2 cells (Muraro et al.,
2016, as assembled by Luecken et al., 2022). `view` holds these programs,
with effects for all 14,528 published genes, restored without refitting; see
[Pancreas programs](pancreas-programs.html) for how the local cache is prepared
and restored. The code runs only when `FLASHIER_UTILS_PANCREAS_DIR` points to
that cache and otherwise shows without output. Upstream data are not
redistributed.

## Rank features

Mean activity per published cell type (scaled to each program's maximum)
shows what the programs describe in the 48 selected CEL-Seq2 cells. `F1` is
active in every type, whereas `F2`, `F4` and `F9` are concentrated in acinar,
beta and ductal cells, respectively.

```{r activity-by-type}
cell_info <- cache$metadata[match(rownames(view$activity),cache$metadata$id),]
scaled <- sweep(view$activity,2,apply(view$activity,2,max),"/")
round(apply(scaled[,c("F1","F2","F4","F9")],2,
            function(a) tapply(a,droplevels(cell_info$celltype),mean)),2)
```

```{r rankings}
largest <- factor_features(view, factor = "F1", select = "largest", n = 6)
largest[, c("feature", "estimate")]

distinctive <- factor_features(view, factor = "F1", select = "distinctive", n = 6)
distinctive[, c("feature", "estimate", "distinctiveness")]

factor_features(view, factor = "F4", select = "distinctive", n = 6)[,
  c("feature", "estimate", "distinctiveness")]
```

`largest` selects the strongest absolute effects.
`distinctive` selects effects that exceed the most extreme same-direction
effect among competing factors, with zero included as a competitor.
A feature can contribute strongly to several factors without distinguishing one.
The table keeps its signed effect in `estimate` and its competitor margin in
`distinctiveness`. For `F1`, `SST` is among the largest effects but not among
the distinctive genes, and every distinctive margin is smaller than its
effect, because other programs also give these genes large effects. `F4`
is the opposite case: the margins for `INS` and `IAPP` stay close to their
effects.

## Plot selected features

```{r plot}
plot_factor_features(view, factors = c("F1","F4"), select = "largest", n = 5)
plot_factor_features(view, factors = c("F1","F4"), select = "distinctive", n = 5)
```

Colors show the signed feature effects. Rankings are descriptive and depend
on the factor scale and sign convention; here effects follow the published
infinity-normalized coordinates, not log-fold changes. Keep the same
representation when comparing factors. See
[Factor scales and measurement units](representation-units.html).

## Sources

The ranking functions adapt `rank_effects` and `compute_le_effects` from
singlecelljamboreeR by Peter Carbonetto and Matthew Stephens.
The plot follows a workflow by David Zemmour.
The data and factors come from Peter Carbonetto's single-cell Jamboree
(CEL-Seq2 cells from Muraro et al., 2016, as assembled by Luecken et al.,
2022); native EBMF methods credit Jason Willwerscheid, Peter Carbonetto,
Wei Wang and Matthew Stephens.
[Original authors and sources](credits.html) lists licenses and citations.

## Competitor comparisons

[Peter Carbonetto's pancreas tutorial](https://stephenslab.github.io/single-cell-jamboree/pancreas_annotate.html)
compares defining effects across programs. Choose `compare_factors` to restrict
competition; NULL retains all-factor competition. Here `F2` and `F9` carry the
exocrine genes of acinar and ductal cells, respectively, so competing only
against `F9` asks which genes separate this pair.
`plot_feature_rankings()` shows untruncated directional differences, separately
from clipped least-extreme margins. See
[worked comparisons and annotations](program-annotation.html).

```{r pancreas-comparisons}
factor_features(view, "F2", n = 5, compare_factors = "F9")[,
  c("feature", "estimate", "distinctiveness", "selection")]
plot_feature_rankings(view, "F2", compare_factors = "F9",
                      direction = "positive", n = 5)
```
