---
title: "Find defining features"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Find defining features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(flashier.utils)
library(Matrix)
set.seed(3)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=6, fig.height=4)
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache_file <- file.path(pancreas_dir,"pancreas-worked-example.rds")
have_pancreas <- nzchar(pancreas_dir) && file.exists(cache_file)
knitr::opts_chunk$set(eval=have_pancreas)
```

Use feature effects to identify measurements that describe a factor.

## Data and objects

**Data.** Nine non-negative matrix factorization (NMF) programs learned from CEL-Seq2 cells (Muraro et al., 2016, as assembled by Luecken et al., 2022), published in [Peter Carbonetto's single-cell Jamboree](https://stephenslab.github.io/single-cell-jamboree/pancreas_annotate.html). Their effects cover 14,528 genes. `view` holds the programs, restored without refitting.

**Get the data.** The package does not ship these data. From a clone of the
[repository](https://github.com/mdmanurung/flashier-utils), run this once; it downloads about 86 MB from a pinned
revision of the Jamboree repository ([details](pancreas-programs.html#prepare-a-local-example-cache)):

```sh
FLASHIER_UTILS_PANCREAS_DIR=~/pancreas-jamboree-example Rscript .github/scripts/prepare-pancreas.R
```

This article reads `pancreas-worked-example.rds`. Set `FLASHIER_UTILS_PANCREAS_DIR` to the same directory before running the
code below; without it the code chunks are skipped. This website was built with the data.

**Objects.** `cache` is the list read from `pancreas-worked-example.rds`: the published programs, 48 reference cells (eight per published cell type), their shifted-log expression and metadata. `view` is the nine-program representation. `restore_programs()` is defined in [restore-programs.R](https://github.com/mdmanurung/flashier-utils/blob/main/vignettes/restore-programs.R) and explained in [Pancreas programs](pancreas-programs.html#restore-published-coordinates-without-refitting).

```{r load-programs}
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache <- readRDS(file.path(pancreas_dir, "pancreas-worked-example.rds"))
source("restore-programs.R")  # from vignettes/ in the repository
view <- restore_programs(cache$reference, cache)
```

## Rank features

Mean activity per published cell type (scaled to each program's maximum)
shows what the programs describe in the 48 selected CEL-Seq2 cells. `F1` is
active in every type, whereas `F2`, `F4` and `F9` are concentrated in acinar,
beta and ductal cells, respectively.

```{r activity-by-type}
cell_info <- cache$metadata[match(rownames(view$activity),cache$metadata$id),]
scaled <- sweep(view$activity,2,apply(view$activity,2,max),"/")
round(apply(scaled[,c("F1","F2","F4","F9")],2,
            function(a) tapply(a,droplevels(cell_info$celltype),mean)),2)
```

```{r rankings}
largest <- factor_features(view, factor = "F1", select = "largest", n = 6)
largest[, c("feature", "estimate")]

distinctive <- factor_features(view, factor = "F1", select = "distinctive", n = 6)
distinctive[, c("feature", "estimate", "distinctiveness")]

factor_features(view, factor = "F4", select = "distinctive", n = 6)[,
  c("feature", "estimate", "distinctiveness")]
```

`largest` selects the strongest absolute effects.
`distinctive` selects effects that exceed the most extreme same-direction
effect among competing factors, with zero included as a competitor.
A feature can contribute strongly to several factors without distinguishing one.
The table keeps its signed effect in `estimate` and its competitor margin in
`distinctiveness`. For `F1`, `SST` is among the largest effects but not among
the distinctive genes, and every distinctive margin is smaller than its
effect, because other programs also give these genes large effects. `F4`
is the opposite case: the margins for `INS` and `IAPP` stay close to their
effects.

## Rank every feature

`factor_features()` returns the top few. `rank_features()` ranks all genes for each
factor, with rank 1 the strongest, and `factor_distinctiveness()` returns the
competitor margin for every gene. Both keep the full table for your own filtering.

```{r full-rankings}
ranks <- rank_features(view, factors = "F4", direction = "positive")
head(ranks[order(ranks$rank), c("feature", "estimate", "rank")], 5)

margins <- factor_distinctiveness(view, direction = "positive")
margins[margins$factor == "F4" & margins$feature %in% c("INS", "IAPP", "SST"),
        c("feature", "estimate", "distinctiveness")]
```

`direction` is `"absolute"`, `"positive"` or `"negative"`. Ties take the average
rank. Use `format = "matrix"` for a feature-by-factor matrix of margins.

## Plot selected features

```{r plot}
plot_factor_features(view, factors = c("F1","F4"), select = "largest", n = 5)
plot_factor_features(view, factors = c("F1","F4"), select = "distinctive", n = 5)
```

Colors show the signed feature effects. Rankings are descriptive and depend
on the factor scale and sign convention; here effects follow the published
infinity-normalized coordinates, not log-fold changes. Keep the same
representation when comparing factors. See
[Factor scales and measurement units](representation-units.html).

## Competitor comparisons

[Peter Carbonetto's pancreas tutorial](https://stephenslab.github.io/single-cell-jamboree/pancreas_annotate.html)
compares defining effects across programs. Choose `compare_factors` to restrict
competition; NULL retains all-factor competition. Here `F2` and `F9` carry the
exocrine genes of acinar and ductal cells, respectively, so competing only
against `F9` asks which genes separate this pair.
`plot_feature_rankings()` shows untruncated directional differences, separately
from clipped least-extreme margins. See
[worked comparisons and annotations](program-annotation.html).

```{r pancreas-comparisons}
factor_features(view, "F2", n = 5, compare_factors = "F9")[,
  c("feature", "estimate", "distinctiveness", "selection")]
plot_feature_rankings(view, "F2", compare_factors = "F9",
                      direction = "positive", n = 5)
```

## Sources

The ranking functions adapt `rank_effects` and `compute_le_effects` from
singlecelljamboreeR by Peter Carbonetto and Matthew Stephens.
The plot follows a workflow by David Zemmour.
The data and factors come from Peter Carbonetto's single-cell Jamboree
(CEL-Seq2 cells from Muraro et al., 2016, as assembled by Luecken et al.,
2022); native EBMF methods credit Jason Willwerscheid, Peter Carbonetto,
Wei Wang and Matthew Stephens.
[Original authors and sources](credits.html) lists licenses and citations.

## Function credits

These are the functions from this package that the article calls, with the people
who wrote the upstream code or method behind each one. Use is this package's
classification and code reuse for the function; `provenance()` returns the full
registry. [Original authors and sources](credits.html) gives the full credits,
citations and licenses.

```{r function-credits, echo=FALSE, eval=TRUE}
code <- knitr::knit_code$get()
used <- unique(unlist(lapply(code[names(code) != "function-credits"], function(x)
  tryCatch(all.names(parse(text = x)), error = function(e) NULL))))
registry <- provenance()
credits <- registry[registry[["function"]] %in% used, ]
credits$use <- paste(credits$classification, credits$code_reuse, sep = " / ")
knitr::kable(credits[, c("function", "authors", "conceptual_origin", "use")],
             col.names = c("Function", "Original authors", "Upstream", "Use"),
             row.names = FALSE)
```
