---
title: "Compare effects across conditions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare effects across conditions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(flashier.utils)
library(Matrix)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5)
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache_file <- file.path(pancreas_dir,"pancreas-cells.rds")
have_pancreas <- nzchar(pancreas_dir) && file.exists(cache_file)
knitr::opts_chunk$set(eval=have_pancreas)
```

Use mashr to shrink comparable factor coefficients across conditions.
Coefficients must describe the same contrast in the same activity representation.

The example uses **real pancreas cells** from
[Peter Carbonetto's single-cell Jamboree](https://stephenslab.github.io/single-cell-jamboree/pancreas.html)
(pinned revision 6cf9aa72): human pancreas cells from the cross-technology integration benchmark of
Luecken et al. (2022) as assembled there; the CEL-Seq2 cells are from Muraro et
al. (2016), and the Jamboree page links the original studies of the other
technologies. The conditions are the four measurement technologies (CEL-Seq2,
SMART-Seq2, Fluidigm C1 and inDrop3), each treated as one cohort, and the shared
contrast is beta minus alpha cells. A local cache holds up to 40 cells per
technology and cell type. It is built outside package builds by the repository
script `.github/scripts/prepare-pancreas.R`, so data chunks run only when
`FLASHIER_UTILS_PANCREAS_DIR` points to that cache, and nothing is downloaded.
The cells are not redistributed with the package. Native EBMF fitting credits
Jason Willwerscheid, Peter Carbonetto, Wei Wang and Matthew Stephens.

## Fit one shared set of factors

Scale each cell to the CEL-Seq2 reference library mean, apply `log1p`, keep the
1,000 genes with the largest variance and fit eight non-negative factors on all
cells with `fit_ebmf()`. One shared fit gives every technology the same factors,
scales and signs, which is what makes their coefficients comparable. These
factors are fitted here; they are not the published Jamboree factors. The gene
selection uses every cell, so it is a global choice.

```{r fit}
cache <- readRDS(cache_file)
metadata <- cache$metadata
metadata$sample <- as.character(metadata$id)
rownames(metadata) <- metadata$sample
metadata$celltype <- droplevels(factor(metadata$celltype))
metadata$tech <- droplevels(factor(metadata$tech))
counts <- cache$counts
stopifnot(identical(rownames(counts), metadata$sample))
Y <- Diagonal(x = cache$reference_library_mean / rowSums(counts)) %*% counts
Y <- log1p(Y)
dimnames(Y) <- dimnames(counts)
variance <- colMeans(Y^2) - colMeans(Y)^2
genes <- names(sort(variance, decreasing = TRUE))[1:1000]
X <- as.matrix(Y[, genes])
fit <- fit_ebmf(X, sample_side = "rows", ebnm_fn = ebnm::ebnm_point_exponential,
                max_factors = 8, backfit = TRUE, nullcheck = FALSE, seed = 1,
                verbose = 0,
                feature_scale = "shifted_log_reference_library_normalized_measurements")
view <- standardize_factors(fit)
dim(view$activity)
```


## Prepare coefficients

Within each technology, regress each factor's activity on cell type using only
alpha and beta cells. Include every eligible factor in every condition,
including nonsignificant rows, and keep the activity representation attached to
the table.

```{r coefficients}
ids <- view$manifest$factor_ids
contrast <- function(technology) {
  cells <- metadata$tech == technology & metadata$celltype %in% c("alpha", "beta")
  group <- factor(metadata$celltype[cells], levels = c("alpha", "beta"))
  do.call(rbind, lapply(ids, function(id) {
    model <- summary(lm(view$activity[cells, id] ~ group))$coefficients["groupbeta", ]
    data.frame(factor = id, condition = technology, term = "beta-minus-alpha",
               estimate = model[["Estimate"]], std.error = model[["Std. Error"]],
               basis_id = view$manifest$basis_id)
  }))
}
coefficients <- do.call(rbind, lapply(levels(metadata$tech), contrast))
attr(coefficients, "analysis_metadata") <- view$manifest
reshape(coefficients[, c("factor", "condition", "estimate")], idvar = "factor",
        timevar = "condition", direction = "wide")
```

Each technology contributes 80 cells to this contrast. The standard errors treat
cells as independent. The cache has no donor identifiers and cells from one donor
are correlated, so these standard errors are optimistic and the coefficients
describe cells, not donors.

Fits learned separately can have different factors, scales and signs.
Their coefficients cannot be pooled directly.

## Shrink effects

```{r shrink}
if (requireNamespace("mashr", quietly = TRUE)) {
  result <- shrink_factor_effects(coefficients, condition = "condition",
                                  term = "beta-minus-alpha", independent_conditions = TRUE)
  result$table[, c("factor", "condition", "estimate", "posterior.sd", "lfsr")]
}
```

The table contains posterior means, posterior standard deviations and local
false sign rates (`lfsr`). `independent_conditions = TRUE` declares independent
estimation errors across conditions. The four technologies contribute different
cells, so this is a reasonable starting point, but they are not independent
replicates of one population, and technology cannot be separated from study and
sample effects in these data. If errors are correlated,
supply their named correlation matrix as `V`.

The shared contrast is concentrated in two factors. `F4`, the alpha factor, is
lower in beta cells and `F5`, the beta factor, is higher, with the same sign and
similar size in all four technologies (`lfsr` near zero). `F6`, one of the
technology factors, is near zero after shrinkage in CEL-Seq2, SMART-Seq2 and
inDrop3 but keeps a positive estimate in Fluidigm C1 only, where it separates
beta from alpha cells. That is a condition-specific effect, not a shared one.
Most remaining rows shrink toward zero with large `lfsr`.

Eight factors in four conditions give mash limited information for learning
patterns of shared effects, and the standard errors above ignore donor
structure. A pooled effect, heterogeneity and cross-condition sharing require
different interpretations. See the function reference for covariance options.

## Sources

The data are from Peter Carbonetto's
[single-cell Jamboree](https://github.com/stephenslab/single-cell-jamboree)
(Muraro et al., 2016; Luecken et al., 2022, as assembled there).
mashr is by Matthew Stephens, Sarah Urbut, Gao Wang, Yuxin Zou and
Peter Carbonetto, with package contributors.
[Original authors and sources](credits.html) lists citations and ashr credits.
