---
title: "Compare effects across conditions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare effects across conditions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use mashr to shrink comparable factor coefficients across conditions.
Coefficients must describe the same contrast in the same activity representation.
The example uses simulated coefficients for two factors in two cohorts.

```{r setup, echo=FALSE}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48), 24, 2), matrix(rnorm(32), 16, 2)) +
  matrix(rnorm(384, sd = 0.2), 24, 16)
dimnames(X) <- list(paste0("s", 1:24), paste0("g", 1:16))
fit <- fit_ebmf(X, sample_side = "rows", seed = 1, max_factors = 2,
                backfit = TRUE, nullcheck = FALSE, verbose = 0,
                feature_scale = "simulated_intensity")
view <- standardize_factors(fit)
```

## Prepare coefficients

Include every eligible factor in every condition, including nonsignificant rows.
Keep the activity representation attached to the table.

```{r coefficients}
ids <- view$manifest$factor_ids
coefficients <- data.frame(
  factor = rep(ids, 2),
  condition = rep(c("cohortA", "cohortB"), each = length(ids)),
  term = "B-minus-A", estimate = seq_len(2 * length(ids)) / 5,
  std.error = 0.5, basis_id = view$manifest$basis_id
)
attr(coefficients, "analysis_metadata") <- view$manifest
```

Fits learned separately can have different factors, scales and signs.
Their coefficients cannot be pooled directly.

## Shrink effects

```{r shrink}
if (requireNamespace("mashr", quietly = TRUE)) {
  result <- shrink_factor_effects(coefficients, condition = "condition",
                                  term = "B-minus-A", independent_conditions = TRUE)
  result$table[, c("factor", "condition", "estimate", "posterior.sd", "lfsr")]
}
```

The table contains posterior means, posterior standard deviations and local
false sign rates (`lfsr`). `independent_conditions = TRUE` declares independent
estimation errors across conditions. If errors are correlated, supply their
named correlation matrix as `V`.

Two factors are enough to illustrate the interface but provide little
information for learning patterns of shared effects.
A pooled effect, heterogeneity and cross-condition sharing require different
interpretations. See the function reference for covariance options.

## Sources

mashr is by Matthew Stephens, Sarah Urbut, Gao Wang, Yuxin Zou and
Peter Carbonetto, with package contributors.
[Original authors and sources](credits.html) lists citations and ashr credits.
