---
title: "Pancreas programs from the single-cell Jamboree"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pancreas programs from the single-cell Jamboree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(flashier.utils)
library(Matrix)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5)
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache_file <- file.path(pancreas_dir,"pancreas-worked-example.rds")
have_pancreas <- nzchar(pancreas_dir) && file.exists(cache_file)
knitr::opts_chunk$set(eval=have_pancreas)
```

This worked example adapts [Peter Carbonetto's pancreas annotation tutorial](https://stephenslab.github.io/single-cell-jamboree/pancreas_annotate.html)
from the single-cell Jamboree. It uses **published CEL-Seq2 NMF and semi-NMF
coordinates**, including the original factors 4–6 comparison, rather than fitting
simulated gene names. CEL-Seq2 data come from Muraro et al. (2016), as assembled
in the pancreas integration benchmark described by Luecken et al. (2022). See
[the source data preparation](https://stephenslab.github.io/single-cell-jamboree/pancreas.html)
and the original studies linked there.

The workflow is independently written using flashier.utils and public flashier
APIs. Peter Carbonetto receives tutorial credit; native EBMF methods credit
Jason Willwerscheid, Peter Carbonetto, Wei Wang and Matthew Stephens. Upstream
workflow code and input data are not redistributed with this package.

## Prepare a local example cache

Download the published objects from a fixed repository revision. This step is
explicit and runs outside package builds. Preparing the cache loads a roughly
640 MiB sparse matrix; subsequent examples use only 48 reference cells and up to
24 query cells, while retaining **all 14,528 reference genes**. The cell subset
contains up to eight cells per acinar, ductal, alpha, beta, delta and gamma group,
chosen by sample ID, and does not change published feature programs.

```{r prepare-cache, eval=FALSE}
# Choose a local archive outside package builds.
pancreas_dir <- path.expand("~/pancreas-jamboree-example")
dir.create(pancreas_dir,recursive=TRUE,showWarnings=FALSE)
revision <- "6cf9aa720cc16dc543e93b36d57bc1a85555860d"
base_url <- paste0("https://raw.githubusercontent.com/stephenslab/",
                   "single-cell-jamboree/",revision,"/")
files <- c("data/pancreas.RData", "output/pancreas_celseq2_factors.RData")
for (file in files) {
  destination <- file.path(pancreas_dir,basename(file))
  if (!file.exists(destination))
    download.file(paste0(base_url,file),destination,mode="wb")
}
original <- new.env(parent=emptyenv())
coordinates <- new.env(parent=emptyenv())
load(file.path(pancreas_dir,"pancreas.RData"),envir=original)
load(file.path(pancreas_dir,"pancreas_celseq2_factors.RData"),envir=coordinates)
metadata <- original$sample_info
stopifnot(identical(rownames(original$counts),as.character(metadata$id)))
reference <- coordinates$fl_nmf_ldf
signed <- coordinates$fl_snmf_ldf
genes <- rownames(reference$F)
stopifnot(identical(rownames(signed$F),genes))
cell_types <- c("acinar","ductal","alpha","beta","delta","gamma")
select_cells <- function(technology,n) {
  unlist(lapply(cell_types,function(type)
    head(sort(as.character(metadata$id[metadata$tech==technology &
                                       metadata$celltype==type])),n)),use.names=FALSE)
}
reference_ids <- select_cells("celseq2",8)
query_ids <- select_cells("smartseq2",4)
# Match the published gene filter before calculating library sizes.
reference_totals <- Matrix::rowSums(original$counts[
  metadata$id[metadata$tech=="celseq2"],genes,drop=FALSE])
reference_library_mean <- mean(reference_totals)
shifted_log <- function(ids) {
  counts <- original$counts[ids,genes,drop=FALSE]
  totals <- Matrix::rowSums(counts)
  stopifnot(all(totals>0))
  Y <- Matrix::Diagonal(x=reference_library_mean/totals) %*% counts
  Y <- log1p(Y)
  dimnames(Y) <- list(ids,genes)
  Y
}
cache <- list(reference=reference,signed=signed,
  reference_expression=shifted_log(reference_ids),
  query_expression=shifted_log(query_ids),
  metadata=metadata[match(c(reference_ids,query_ids),metadata$id),],
  revision=revision,reference_library_mean=reference_library_mean,
  source_files=files,
  source_sha256=vapply(basename(files),function(file)
    digest::digest(file=file.path(pancreas_dir,file),algo="sha256"),character(1)))
saveRDS(cache,file.path(pancreas_dir,"pancreas-worked-example.rds"))
```

The repository script
[`.github/scripts/prepare-pancreas.R`](https://github.com/mdmanurung/flashier-utils/blob/main/.github/scripts/prepare-pancreas.R)
runs this code and also writes a second file, `pancreas-cells.rds`, with raw counts
for up to 40 cells per technology and cell type (896 cells from CEL-Seq2,
SMART-Seq2, Fluidigm C1 and inDrop3). Other tutorials read it. From a clone of the
repository, run it once (about 86 MB are downloaded):

```sh
FLASHIER_UTILS_PANCREAS_DIR=~/pancreas-jamboree-example Rscript .github/scripts/prepare-pancreas.R
```

Set `FLASHIER_UTILS_PANCREAS_DIR` to this directory before running the code below
or rendering any tutorial that uses these data. Package builds skip
data-dependent execution when the cache is absent and never perform network
downloads. This website was built with the data.

```{r load-cache}
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache <- readRDS(file.path(pancreas_dir, "pancreas-worked-example.rds"))
cache$revision
dim(cache$reference_expression)
dim(cache$query_expression)
```

## Restore published coordinates without refitting

The upstream file contains `L`, `D`, and `F` rather than serialized native fits.
Initialize a native object with the measured reference matrix, restore the
published posterior means, and fix the feature side. flashier fixes one mode per
factor; the activities are restored unchanged because no backfit follows. Absorbing `D` into activities keeps
feature coordinates equal to the source tutorial's infinity-normalized `F`.
This restores means only; it does not restore upstream posterior uncertainties or
hyperparameters. No reference model is refitted.

```{r restore-programs, file="restore-programs.R"}
```

```{r restore-calls}
view <- restore_programs(cache$reference,cache)
signed_view <- restore_programs(cache$signed,cache,signed_effects=TRUE)
# The input matrix uses shifted log measurements. The unmodified tutorial F
# has its own fixed coordinate normalization; it is not a log-fold change.
```

## Borrow the original factors 4–6 comparison

Carbonetto's worked comparison focuses on factors 4, 5 and 6 and chooses eight
genes per factor, competing within that same set. Here `F4`–`F6` preserve those
source column indices. Compare largest effects, clipped distinctive margins,
and untruncated differences in separate displays.

```{r pancreas-annotation}
islet_factors <- paste0("F",4:6)
factor_features(view,"F4",select="both",direction="positive",n=8,
                compare_factors=islet_factors)[,c("feature","estimate","selection")]
plot_factor_features(view,islet_factors,n=8,direction="positive")
plot_factor_features(view,islet_factors,select="distinctive",n=8,
                     direction="positive",compare_factors=islet_factors)
plot_feature_rankings(view,"F4",n=8,compare_factors=islet_factors)
```

Check `INS` and `IAPP` as beta-associated evidence and `GCG` as alpha-associated
evidence, following the source tutorial. These are evidence to inspect alongside
published cell metadata, not automatic cell-type assignments. In particular,
shared endocrine genes and factors containing multiple cell types warrant
cautious user-authored labels.

```{r pancreas-explicit-markers}
markers <- c("INS","IAPP","GCG","SST","PPY","ISL1","CPE")
plot_factor_features(view,islet_factors,features=markers,
                     annotation=c(F4="INS/IAPP evidence",F5="GCG evidence",
                                  F6="Inspect SST/PPY overlap"))
# Semi-NMF effects preserve both signs; choose competing factors explicitly.
plot_factor_features(signed_view,paste0("F",4:7),direction="negative",n=5)
```

## Shared-gene networks and descriptive diagnostics

```{r pancreas-networks}
network <- factor_network(view,factors=islet_factors,n=5)
network$edges[,c("factor","feature","estimate","weight","sign")]
if(requireNamespace("igraph",quietly=TRUE) && requireNamespace("ggraph",quietly=TRUE))
  plot_factor_network(network,highlight="F4",gene_labels="all")
factor_diagnostics(view,factors=islet_factors)$factors
reference_metadata <- cache$metadata[match(rownames(view$activity),cache$metadata$id),]
reference_metadata$sample <- as.character(reference_metadata$id)
if(requireNamespace("pROC",quietly=TRUE))
  factor_specificity(view,reference_metadata,"celltype",groups=c("alpha","beta"),
                     factors=islet_factors)
```

Prevalence and AUC describe the **selected 48 CEL-Seq2 cells**, not the full
published 2,285-cell population. Feature selection uses every reference gene.
This selected-cell demonstration is not a calibrated cell classifier.

## Transfer the reference programs to another technology

SMART-Seq2 query cells were not among the CEL-Seq2 cells used for the reference
programs. They use the frozen CEL-Seq2 library-size mean, the same reference gene
set and `log1p` transform. An equal mathematical transform does not establish
biological equivalence across technologies: inspect reconstruction and metadata
rather than treating transferred activities as validated annotations. The
benchmark's entries are not uniformly raw counts; retain the input description
from the source dataset.

```{r pancreas-transfer}
transfer <- project_factor_activity(view,cache$query_expression,"rows",
  feature_scale="shifted_log_reference_library_normalized_measurements",
  preprocessing=list(scope="external_reference",revision=cache$revision,
    transformation="log1p(filtered measurement / library size * reference mean)",
    reference_library_mean=cache$reference_library_mean))
stopifnot(identical(transfer$representation$effects,view$effects),
          identical(transfer$representation$offset,view$offset))
transfer$reconstruction
transfer$native_iteration_diagnostics
query_metadata <- cache$metadata[match(rownames(transfer$representation$activity),
                                      cache$metadata$id),]
query_metadata$sample <- as.character(query_metadata$id)
plot_factor_activity(transfer$representation,query_metadata,group="celltype",
                     factors=islet_factors,style="distribution")
```

Transferred activity distributions are descriptive. Hyperparameters and noise may
recalibrate on query observations; fixed feature programs and factor IDs remain
unchanged. Reference-program transfer receives conceptual credit to STARCAT
(Dylan Kotliar and Michelle Curtis), while estimation uses native flashier.

For protein contrasts, mixture thresholds and gates, see [protein interpretation](protein-interpretation.html)
and the [ImmGen Figure 7 source workflow](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/Figure7.R).
This RNA-only pancreas example does not invent protein measurements.

## Function credits

These are the functions from this package that the article calls, with the people
who wrote the upstream code or method behind each one. Use is this package's
classification and code reuse for the function; `provenance()` returns the full
registry. [Original authors and sources](credits.html) gives the full credits,
citations and licenses.

```{r function-credits, echo=FALSE, eval=TRUE}
code <- knitr::knit_code$get()
used <- unique(unlist(lapply(code[names(code) != "function-credits"], function(x)
  tryCatch(all.names(parse(text = x)), error = function(e) NULL))))
registry <- provenance()
credits <- registry[registry[["function"]] %in% used, ]
credits$use <- paste(credits$classification, credits$code_reuse, sep = " / ")
knitr::kable(credits[, c("function", "authors", "conceptual_origin", "use")],
             col.names = c("Function", "Original authors", "Upstream", "Use"),
             row.names = FALSE)
```
