---
title: "Program networks and evidence for annotations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Program networks and evidence for annotations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(flashier.utils)
library(Matrix)
set.seed(9)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5)
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache_file <- file.path(pancreas_dir,"pancreas-worked-example.rds")
have_pancreas <- nzchar(pancreas_dir) && file.exists(cache_file)
knitr::opts_chunk$set(eval=have_pancreas)
```

```{r load-programs, include=FALSE}
# Restore published factors as in the pancreas-programs article (means only).
cache <- readRDS(cache_file)
source("restore-programs.R", local=knitr::knit_global())
view <- restore_programs(cache$reference)
signed_view <- restore_programs(cache$signed,signed_effects=TRUE)
factors <- colnames(view$activity)
```

These examples use **published human-pancreas programs** from Peter Carbonetto's
[single-cell Jamboree](https://stephenslab.github.io/single-cell-jamboree/pancreas_annotate.html):
nine NMF programs (`view`) and nine semi-NMF programs with signed gene effects
(`signed_view`), both learned from CEL-Seq2 cells (Muraro et al., 2016, as
assembled by Luecken et al., 2022) and restored without refitting as described in
[Pancreas programs](pancreas-programs.html). The code runs only when
`FLASHIER_UTILS_PANCREAS_DIR` points to the local cache, and otherwise shows
without output. Upstream data are not redistributed.

The workflows are inspired by ImmGen Figures 1c and 5b and the pancreas
annotation tutorial.
ImmGen workflow repository credit: [Ziang Zhang / AgueroZZ](https://github.com/AgueroZZ/immgenT-GP-analysis);
individual script authors remain unresolved. No upstream workflow source was copied.
Existing ranking helpers credit Peter Carbonetto and Matthew Stephens; native EBMF
methods credit Jason Willwerscheid, Peter Carbonetto, Wei Wang and Matthew Stephens.

## Signed programs and shared genes

```{r network}
islet <- c("F1","F3","F4","F8")
network <- factor_network(signed_view, factors=islet, direction="both", n=4)
network$edges[,c("factor","feature","estimate","sign")]
if (requireNamespace("igraph",quietly=TRUE) && requireNamespace("ggraph",quietly=TRUE)) {
  plot_factor_network(network, highlight="F4", gene_labels="all", seed=42)
}
```

Factor IDs in `signed_view` index the semi-NMF programs and need not match the
NMF IDs in `view`. In these four programs the top genes of both signs overlap:
`GCG` is a positive gene of `F1` and `F3` and a negative gene of `F4` and `F8`,
and `INS` is positive in `F4` and negative in `F3`. Shared neighbors,
not sign agreement, create the links, so read the edge sign when interpreting them.

Each factor and gene has a separate internal ID, including when a gene name
matches a factor name. Shared gene neighbors connect programs. Per-factor maximum
absolute scaling affects display and selection only; edge tables retain original
signed effects and the source basis. `layout="stress"` uses the ggraph engine;
`layout="fr"` uses igraph. A scoped seed restores caller RNG state.

## Compare against a chosen set of programs

`F3` and `F7` are two non-epithelial programs whose top genes overlap
(`SPARC`, `IGFBP7`). Restricting the competition to `F3` shows which `F7` genes
remain after the shared ones are accounted for.

```{r comparisons}
pool <- "F3"
factor_features(view, "F7", n=3, compare_factors=pool)[,
  c("feature","estimate","distinctiveness","selection")]
plot_feature_rankings(view, "F7", compare_factors=pool, direction="positive", n=4)
```

The scatter shows the effect and its **untruncated** difference from the largest
positive-direction competitor (or smallest negative-direction competitor).
It labels the union of top effect and top difference candidates. Clipped
least-extreme margins instead include zero in competition and clip opposing
margins to zero. `compare_factors=NULL` retains competition against every factor;
the target always excludes itself. An empty competitor set uses zero.

## Record user-authored labels

Collect evidence first: the highest-weighted distinctive gene of each program and
the published cell type with the highest mean scaled activity among the 48
selected cells.

```{r evidence}
cell_info <- cache$metadata[match(rownames(view$activity),cache$metadata$id),]
scaled <- sweep(view$activity,2,apply(view$activity,2,max),"/")
mean_activity <- apply(scaled,2,function(a) tapply(a,droplevels(cell_info$celltype),mean))
evidence <- data.frame(factor=factors,
  distinctive_genes=vapply(factors,function(f) paste(
    factor_features(view,f,select="distinctive",n=3)$feature,collapse=", "),""),
  top_cell_type=rownames(mean_activity)[apply(mean_activity,2,which.max)],
  top_mean_activity=round(apply(mean_activity,2,max),2))
evidence
```

Labels below are written by the analyst from this evidence; a question mark marks
a candidate. `F1` is active in every cell type; `F6` has `PPY` among its distinctive genes and, in the
activity table, gamma cells also carry activity. Programs with weak activity in these
48 cells, and no cell type of that kind in the selection, stay unresolved.

```{r annotations, fig.width=9}
labels <- c(F1="Broad", F2="Acinar?", F3="Unresolved", F4="Beta?", F5="Alpha?",
            F6="SST/PPY?", F7="Unresolved", F8="Unresolved", F9="Ductal?")
stopifnot(identical(names(labels),factors))
plot_factor_features(view, factors, direction="positive", n=3, annotation=labels)
shown <- c("F2","F4","F5","F6","F9")
plot_factor_features(view, shown,
                     features=c("REG1A","INS","GCG","SST"),
                     display_normalization="max_abs", transpose=TRUE,
                     annotation=labels[shown])
summary <- summarize_factors(view, n=3, annotations=labels)
summary[,c("factor","active_proportion","active_features","annotation")]
```

The heatmap defaults to matched feature-effect units. A log-fold-change label is
appropriate only when the supplied input and representation have those units;
these published effects are infinity-normalized factor coordinates.
The annotation strip records evidence or labels supplied by the analyst. It does
not assign cell types, and the 48-cell summary is not a calibrated classifier.
See [credits](credits.html) and `provenance(result=network)`.

## Published-data example

See [Pancreas programs from the single-cell Jamboree](pancreas-programs.html)
for the full worked comparison of factors 4-6, diagnostics and transfer to
SMART-Seq2 cells. This article uses the same CEL-Seq2 programs (credit: Peter
Carbonetto's single-cell Jamboree; Muraro et al., 2016; Luecken et al., 2022) to
teach network and annotation functions.
