---
title: "Descriptive program diagnostics and specificity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Descriptive program diagnostics and specificity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(flashier.utils)
library(Matrix)
set.seed(9)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5)
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache_file <- file.path(pancreas_dir,"pancreas-worked-example.rds")
have_pancreas <- nzchar(pancreas_dir) && file.exists(cache_file)
knitr::opts_chunk$set(eval=have_pancreas)
```

```{r load-programs, include=FALSE}
# Restore published factors as in the pancreas-programs article (means only).
cache <- readRDS(cache_file)
restore_programs <- function(ldf,signed_effects=FALSE) {
  Y <- cache$reference_expression
  factor_ids <- paste0("F",seq_len(ncol(ldf$F)))
  A <- sweep(ldf$L[rownames(Y),,drop=FALSE],2,ldf$D,"*")
  B <- ldf$F
  colnames(A) <- colnames(B) <- factor_ids
  priors <- list(ebnm::ebnm_point_exponential,
                 if(signed_effects) ebnm::ebnm_point_normal
                 else ebnm::ebnm_point_exponential)
  object <- fit_ebmf(Y,"rows",ebnm_fn=priors,max_factors=0,var_type=2,seed=1,
    feature_scale="shifted_log_reference_library_normalized_measurements",verbose=0)
  metadata <- attr(object,"flashbridge_metadata")
  metadata$factor_ids <- factor_ids
  metadata$preprocessing <- list(scope="external_reference",revision=cache$revision,
    transformation="log1p(filtered measurement / library size * reference mean)",
    reference_library_mean=cache$reference_library_mean,
    source_sha256=cache$source_sha256)
  object <- flashier::flash_factors_init(object,list(A,B),ebnm_fn=priors)
  object <- flashier::flash_factors_fix(object,seq_along(factor_ids),"factors")
  attr(object,"flashbridge_metadata") <- metadata
  standardize_factors(object,sample_side="rows",scaling="raw",
    prior_support=c(activity="nonnegative",
                    effects=if(signed_effects) "signed" else "nonnegative"))
}
view <- restore_programs(cache$reference)
factors <- colnames(view$activity)
```

This independently written adaptation covers the computational
[ImmGen Figure 2 workflow](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/Figure2.R)
(repository author Ziang Zhang; per-script authors unresolved). It describes
programs in the declared population and does not establish biological replication.

The examples use **published human-pancreas programs** from Peter Carbonetto's
[single-cell Jamboree](https://stephenslab.github.io/single-cell-jamboree/pancreas_annotate.html):
nine NMF programs learned from CEL-Seq2 cells (Muraro et al., 2016, as assembled
by Luecken et al., 2022), restored as `view` without refitting; see
[Pancreas programs](pancreas-programs.html). The population here is **48 selected
CEL-Seq2 cells**, eight from each of six published cell types, with the published
cell-type labels as metadata. The code runs only when `FLASHIER_UTILS_PANCREAS_DIR`
points to the local cache and otherwise shows without output; upstream data are not
redistributed. Native EBMF methods credit Jason Willwerscheid, Peter Carbonetto,
Wei Wang and Matthew Stephens.

```{r cells}
Y <- cache$reference_expression
metadata <- cache$metadata[match(rownames(view$activity),cache$metadata$id),]
metadata$sample <- as.character(metadata$id)
metadata$celltype <- as.character(metadata$celltype)
table(metadata$celltype)
```

## Signature weights and expression

```{r signature}
plot_factor_signature(view, "F4", Matrix::colMeans(Y),
  expression_scale="mean shifted-log expression, 48 CEL-Seq2 cells", n_label=8)
```

The vertical axis is mean expression in these 48 cells, never significance.
For `F4`, the labelled genes (`INS`, `IAPP`, `SLC30A8` among them) have high
weights but only moderate mean expression, as expected for beta-cell genes
averaged over 48 cells of which eight are beta cells. Weights are divided by each
program's maximum absolute weight. Labels require
absolute displayed weight at least 0.1, with ties ordered by feature ID.

## Declare thresholds once

```{r diagnostics}
diagnostics <- factor_diagnostics(view)
diagnostics$factors
head(diagnostics$samples)
ins <- setNames(as.numeric(Y[,"INS"]),rownames(Y))
plots <- plot_factor_diagnostics(view, metadata, "celltype", factor="F4",
                                 marker=ins, max_samples=40)
plots$prevalence
plots$features
plots$programs_per_cell
plots$groups
plots$marker
if (requireNamespace("patchwork",quietly=TRUE))
  plots$prevalence + plots$features + patchwork::plot_layout(ncol=2)
```

In this selection `F1` is active in all 48 cells, `F7` and `F8` in only two, and
cells carry a median of two (alpha) to 4.5 (beta) active programs.
The selected cells include no endothelial or macrophage cells.

Defaults count positive max-absolute-scaled activity strictly above 0.1 and
absolute max-absolute-scaled feature effects strictly above 0.25. Zero columns
remain zero. `activity_direction="negative"` counts activity below -0.1;
`"absolute"` counts either sign. All counts and marker correlations use the full
population (here, `INS` shifted-log expression against `F4` activity);
`max_samples` only reduces scatter rendering.

## One-versus-rest descriptive classification

```{r specificity}
have_proc <- requireNamespace("pROC",quietly=TRUE)
if (have_proc)
  factor_specificity(view, metadata, "celltype", groups=c("beta","acinar"),
                     factors=c("F2","F4"))
# F1 is active in every cell type; acinar cells have the lowest mean F1 activity,
# so ask whether low activity identifies them.
if (have_proc)
  factor_specificity(view, metadata, "celltype", groups="acinar", factors="F1",
                     direction="negative", threshold_method="youden")
```

pROC authors Xavier Robin and collaborators supply the native AUC and threshold
engine. Direction is explicit: positive means higher activity predicts the target
class, negative means lower activity predicts it. The default chooses the closest
top-left optimum. Tied optima use the smallest finite threshold and record their
count. Constants have AUC 0.5 with no threshold; absent classes are unavailable.
Observation counts describe the 48 cells. The metadata carry no donor or other
biological-unit identifier, so `unit_col` is not used and nothing here is
independent of the individual from whom cells were taken. No cell-level
inferential p-values are computed. These are descriptions of a small selected
subset, not a calibrated classifier. See [credits](credits.html) for dependency
authors.

## Published-data example

See [Pancreas programs from the single-cell Jamboree](pancreas-programs.html)
for the full worked comparison of factors 4-6 and transfer to SMART-Seq2 cells.
This article uses the same CEL-Seq2 programs (credit: Peter Carbonetto's
single-cell Jamboree; Muraro et al., 2016; Luecken et al., 2022) to teach the
diagnostic functions.
