---
title: "Protein interpretation and explicit marker gates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Protein interpretation and explicit marker gates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(flashier.utils)
set.seed(9)
A <- matrix(rnorm(160), 80, 2)
B <- matrix(rnorm(40), 20, 2)
X <- tcrossprod(A, B) + matrix(rnorm(1600, sd=0.2), 80, 20)
dimnames(X) <- list(paste0("cell", 1:80), paste0("gene", 1:20))
fit <- fit_ebmf(X, "rows", max_factors=2, nullcheck=FALSE, backfit=TRUE,
                ebnm_fn=list(ebnm::ebnm_normal, ebnm::ebnm_point_laplace),
                feature_scale="simulated_centered_expression", seed=10, verbose=0)
view <- standardize_factors(fit, scaling="raw")
factors <- colnames(view$activity)
stopifnot(length(factors) == 2)
metadata <- data.frame(sample=rownames(X), lineage=rep(c("A","B"),each=40),
                       donor=rep(paste0("donor",1:20),each=4))
```

This independent adaptation follows computational
[ImmGen Figure 7](https://github.com/AgueroZZ/immgenT-GP-analysis/blob/main/script/Figure7.R)
(repository author Ziang Zhang; per-script authors unresolved). The original
schematic is hand drawn. These worked examples use simulated proteins and gates,
not biological labels. Existing contrast workflows credit David Zemmour.

```{r proteins}
proteins <- cbind(CD1=view$activity[,1]+rnorm(80,sd=0.3),
                  CD2=view$activity[,2]+rnorm(80,sd=0.3))
rownames(proteins) <- rownames(X)
correlations <- correlate_factor_markers(view, proteins, method="spearman")
correlations
strip <- correlations[correlations$marker=="CD1",]
annotation <- setNames(round(strip$correlation,2),strip$factor)
plot_factor_features(view, factors, direction="positive", n=4, annotation=annotation)
```

Correlations use aligned sample IDs and the full declared population; they are
descriptive and contain no cell-level p-values. For marker-positive versus
marker-negative comparisons within lineages, supply corresponding metadata to
`compare_factor_groups()` before comparing effects:

```{r contrasts}
metadata$marker <- ifelse(proteins[,"CD1"]>0,"positive","negative")
# Conditional lineage-specific group labels on the same source basis.
metadata$gate <- paste(metadata$lineage,metadata$marker)
a <- compare_factor_groups(view,metadata,"gate",c("A positive","A negative"))
b <- compare_factor_groups(view,metadata,"gate",c("B positive","B negative"))
plot_factor_contrasts(a,b,n_label=2)
```

## Candidate mixture thresholds require review

```{r mixtures}
mixture_protein <- matrix(c(rnorm(40,2,0.1),rnorm(40,5,0.1)),80,1,
                         dimnames=list(rownames(X),"CDmix"))
if (requireNamespace("mclust",quietly=TRUE)) {
  candidates <- estimate_marker_thresholds(mixture_protein, thresholds=c(CDmix=3))
  candidates$thresholds
  candidates$mixtures$CDmix
  plot_marker_thresholds(candidates,mixture_protein)
}
```

Native mclust defaults fit one- and two-component Gaussian models to values above
0.5, requiring at least 50 observations and using scoped seed 42. Candidate is the
largest eligible value assigned to the lower-mean component. Outputs retain sizes,
parameters, warnings, exclusions, pooled-SD separation and BIC difference.
Failures, empty components, separation below one SD or no BIC support require
manual review. Supplied thresholds remain authoritative. Engine credit: Luca
Scrucca, Chris Fraley, Adrian E. Raftery and collaborators; see [credits](credits.html).

## Compare equal-size populations

```{r gates}
alignment <- factor_gate_alignment(view, proteins, factors[1], positive="CD1",
                                   negative="CD2", thresholds=c(CD1=0,CD2=0))
alignment$summary
embedding <- view$activity[,1:2,drop=FALSE]
plot_factor_gates(alignment,embedding)
plot_factor_gates(alignment,embedding,display="density",bins=10)
```

Positive gates use `>` and negative gates use `<=`. Thresholds are explicit;
missing markers or thresholds fail. The activity population contains exactly as
many cells as the gate, with activity ties broken by sample ID. Empty gates report
unavailable overlap. Both maps use the same full population and geometry.
