---
title: "Code origins and runtime versions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Code origins and runtime versions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48),24,2),matrix(rnorm(32),16,2)) + matrix(rnorm(384,sd=0.2),24,16)
dimnames(X) <- list(paste0("s",1:24),paste0("g",1:16))
fit <- fit_ebmf(X,"rows",seed=1,max_factors=2,verbose=0,backfit=TRUE,nullcheck=FALSE,feature_scale="arcsinh_intensity")
view <- standardize_factors(fit)
metadata <- data.frame(sample=rownames(X),donor=rep(paste0("d",1:12),each=2),group=rep(c("A","B"),each=12),visit=rep(0:1,12))
```

Source inspection, numerical parity, scientific calibration and runtime installation evidence are separate records. Audited source SHAs are never presented as installed SHAs when the installed package lacks one.
```{r}
provenance(c("standardize_factors","backproject_contrast","factor_distinctiveness"))
```
The package contains two modified MIT feature-ranking helpers with the complete source notice. Other adapters call public native functions or independently implement credited glue. The installed NOTICE and provenance CSV list sources and changes. Package-owned code is maintained by Mikhael Manurung; GitHub distribution is checked through cross-platform CI.


## Source credits

Original function and package authors are named in each function manual and in
[Original authors and sources](credits.html). The installed `provenance()` registry
records source pins, authors, licenses and reuse. Please cite the upstream methods
used in your analysis; `citation("flashier")` and `citation("ebnm")` give core citations.
