---
title: "Find sources and cite methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Find sources and cite methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Cite flashier and ebnm for factorization, and cite the optional methods used
in your analysis. The citation matches the installed package version.

```{r citations}
library(flashier.utils)
citation("flashier")
citation("ebnm")
```

## Find a function's source

```{r sources}
sources <- provenance(c("standardize_factors", "backproject_contrast",
                        "factor_distinctiveness"))
sources[, c("function", "authors", "classification", "code_reuse")]
```

The registry distinguishes wrappers, modified source and conceptual workflow
adaptations. The two modified singlecelljamboreeR ranking helpers are by
Peter Carbonetto and Matthew Stephens. Plotting and reconstruction workflows
credit David Zemmour and the Zemmour Lab. Mikhael Manurung maintains this package.

[Original authors and sources](credits.html) lists all credits and citations.
The installed `NOTICE` and license files retain upstream notices.

## Record package versions

```{r versions}
attr(sources, "dependencies")
```

Save this information with analysis results to record installed versions.
An audited source commit identifies the reviewed source. The runtime commit
identifies the installed package when that information is available.

## Function credits

These are the functions from this package that the article calls, with the people
who wrote the upstream code or method behind each one. Use is this package's
classification and code reuse for the function; `provenance()` returns the full
registry. [Original authors and sources](credits.html) gives the full credits,
citations and licenses.

```{r function-credits, echo=FALSE, eval=TRUE}
code <- knitr::knit_code$get()
used <- unique(unlist(lapply(code[names(code) != "function-credits"], function(x)
  tryCatch(all.names(parse(text = x)), error = function(e) NULL))))
registry <- provenance()
credits <- registry[registry[["function"]] %in% used, ]
credits$use <- paste(credits$classification, credits$code_reuse, sep = " / ")
knitr::kable(credits[, c("function", "authors", "conceptual_origin", "use")],
             col.names = c("Function", "Original authors", "Upstream", "Use"),
             row.names = FALSE)
```
