---
title: "Factor scales and measurement units"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factor scales and measurement units}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Keep activity and feature effects from the same representation.
Their product reconstructs the fitted matrix with samples in rows.
This example fits the GTEx z-score matrix that ships with flashier. Its
columns are tissues, so `sample_side = "columns"`.

```{r setup}
library(flashier.utils)
data("gtex", package = "flashier")
fit <- fit_ebmf(gtex, sample_side = "columns", max_factors = 2,
                backfit = TRUE, seed = 1, verbose = 0,
                feature_scale = "association_z_score")
view <- standardize_factors(fit)
```

## Extract matching matrices

```{r matrices}
A <- factor_activity(view)
B <- factor_effects(view)
dim(A)
dim(B)
```

Rows of `A` are samples (the 44 tissues). Rows of `B` are features (the 1000
SNP-gene pairs). Their columns are matching factors. `tcrossprod(A, B)` gives
the represented sample-by-feature matrix, which is the transpose of `gtex`.

## Choose a scale

By default, `standardize_factors()` normalizes the factors with flashier's `ldf()`
and includes the factor magnitude in the feature effects. `scaling = "raw"` uses the native
posterior means. These choices preserve the reconstruction.

```{r scales}
raw <- standardize_factors(fit, scaling = "raw")
max(abs(tcrossprod(raw$activity, raw$effects) - tcrossprod(A, B)))
```

The difference is at the level of numerical rounding.

Changing the scale or sign changes the meaning of an activity coefficient.
Fit your activity model on the representation you will use for
[backprojection](backprojection.html). Save `view` with the model results.

## Retain measurement units

```{r units}
view$manifest$feature_scale
```

`feature_scale` records the units supplied to `fit_ebmf()`; here they are
GTEx association z-scores.
The package does not normalize counts, invert nonlinear transformations or
convert estimates to log2 fold changes. Centering or scaling activity outside
the package requires matching changes to feature effects and any feature offset.

## Sources

The GTEx data ship with flashier (Jason Willwerscheid, Peter Carbonetto, Wei
Wang and Matthew Stephens) and derive from the GTEx project; see `?flashier::gtex`.

The representation uses flashier's LDF interface by Jason Willwerscheid,
Peter Carbonetto, Wei Wang and Matthew Stephens.
[Original authors and sources](credits.html) lists method citations.
