---
title: "Factor scales and measurement units"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factor scales and measurement units}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Keep activity and feature effects from the same representation.
Their product reconstructs the fitted matrix with samples in rows.
This example uses a simulated fit.

```{r setup, echo=FALSE}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48), 24, 2), matrix(rnorm(32), 16, 2)) +
  matrix(rnorm(384, sd = 0.2), 24, 16)
dimnames(X) <- list(paste0("s", 1:24), paste0("g", 1:16))
fit <- fit_ebmf(X, sample_side = "rows", seed = 1, max_factors = 2,
                backfit = TRUE, nullcheck = FALSE, verbose = 0,
                feature_scale = "simulated_intensity")
view <- standardize_factors(fit)
```

## Extract matching matrices

```{r matrices}
A <- factor_activity(view)
B <- factor_effects(view)
dim(A)
dim(B)
```

Rows of `A` are samples. Rows of `B` are features. Their columns are matching
factors. `tcrossprod(A, B)` gives the represented sample-by-feature matrix.

## Choose a scale

By default, `standardize_factors()` normalizes the factors with flashier's `ldf()`
and includes the factor magnitude in the feature effects. `scaling = "raw"` uses the native
posterior means. These choices preserve the reconstruction.

```{r scales}
raw <- standardize_factors(fit, scaling = "raw")
max(abs(tcrossprod(raw$activity, raw$effects) - tcrossprod(A, B)))
```

Changing the scale or sign changes the meaning of an activity coefficient.
Fit your activity model on the representation you will use for
[backprojection](backprojection.html). Save `view` with the model results.

## Retain measurement units

```{r units}
view$manifest$feature_scale
```

`feature_scale` records the units supplied to `fit_ebmf()`.
The package does not normalize counts, invert nonlinear transformations or
convert estimates to log2 fold changes. Centering or scaling activity outside
the package requires matching changes to feature effects and any feature offset.

## Sources

The representation uses flashier's LDF interface by Jason Willwerscheid,
Peter Carbonetto, Wei Wang and Matthew Stephens.
[Original authors and sources](credits.html) lists method citations.
