---
title: "Named single-cell reductions and display percentiles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Named single-cell reductions and display percentiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48),24,2),matrix(rnorm(32),16,2)) + matrix(rnorm(384,sd=0.2),24,16)
dimnames(X) <- list(paste0("s",1:24),paste0("g",1:16))
fit <- fit_ebmf(X,"rows",seed=1,max_factors=2,verbose=0,backfit=TRUE,nullcheck=FALSE,feature_scale="arcsinh_intensity")
view <- standardize_factors(fit)
metadata <- data.frame(sample=rownames(X),donor=rep(paste0("d",1:12),each=2),group=rep(c("A","B"),each=12),visit=rep(0:1,12))
```

The example treats each simulated row as a cell for adapter mechanics only. It makes no donor-inference or private ImmGenT reproduction claim. The original activity remains the inference coordinate; percentiles are display-only.
```{r}
percentile <- factor_activity_percentile(view,reference_samples=rownames(X)[1:12])
head(percentile)
if (requireNamespace("SeuratObject",quietly=TRUE) && requireNamespace("uwot",quietly=TRUE)) {
  counts <- Matrix::Matrix(matrix(1,16,24,dimnames=list(colnames(X),rownames(X))),sparse=TRUE)
  object <- SeuratObject::CreateSeuratObject(counts)
  object <- add_factors_to_seurat(object,view)
  object <- embed_factors(object,seed=15,n_neighbors=4,n_epochs=20,init="random")
  plot_factor_embedding(object,"ebmfumap","F1",display="percentile",max_samples=20,seed=12)
}
```
Exactly matching cell IDs are required; donor scores are never implicitly expanded. Loadings are omitted if assay feature IDs differ. UMAP is exploratory geometry, not batch correction or program replication. Native ECDF ties and named reference IDs are retained. The million-cell benchmark measures bounded display construction, not fitting or UMAP at that size.


## Source credits

Original function and package authors are named in each function manual and in
[Original authors and sources](credits.html). The installed `provenance()` registry
records source pins, authors, licenses and reuse. Please cite the upstream methods
used in your analysis; `citation("flashier")` and `citation("ebnm")` give core citations.
