---
title: "Display factors in Seurat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Display factors in Seurat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(flashier.utils)
library(Matrix)
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5)
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache_file <- file.path(pancreas_dir,"pancreas-cells.rds")
have_pancreas <- nzchar(pancreas_dir) && file.exists(cache_file)
knitr::opts_chunk$set(eval=have_pancreas)
```

Store factor activities as a Seurat reduction and plot an embedding with
uniform manifold approximation and projection (UMAP). Install SeuratObject and
uwot to run this example.

## Data and objects

**Data.** Human pancreas cells from the cross-technology integration benchmark of Luecken et al. (2022) as assembled in [Peter Carbonetto's single-cell Jamboree](https://stephenslab.github.io/single-cell-jamboree/pancreas.html) (pinned revision 6cf9aa72). The CEL-Seq2 cells are from Muraro et al. (2016); the Jamboree page links the original studies of the other technologies. The cache holds up to 40 cells per technology (CEL-Seq2, SMART-Seq2, Fluidigm C1 and inDrop3) and cell type: 896 cells. Native EBMF fitting credits Jason Willwerscheid, Peter Carbonetto, Wei Wang and Matthew Stephens.

**Get the data.** The package does not ship these data. From a clone of the
[repository](https://github.com/mdmanurung/flashier-utils), run this once; it downloads about 86 MB from a pinned
revision of the Jamboree repository ([details](pancreas-programs.html#prepare-a-local-example-cache)):

```sh
FLASHIER_UTILS_PANCREAS_DIR=~/pancreas-jamboree-example Rscript .github/scripts/prepare-pancreas.R
```

This article reads `pancreas-cells.rds`. Set `FLASHIER_UTILS_PANCREAS_DIR` to the same directory before running the
code below; without it the code chunks are skipped. This website was built with the data.

**Objects.** `cache` is the list read from `pancreas-cells.rds`: raw counts, cell metadata and the CEL-Seq2 reference library mean. `fit` and `view` are created in the next section; the Seurat object is created from them below.

## Fit factors on shifted-log expression

Scale each cell to the CEL-Seq2 reference library mean, apply `log1p` and keep
the 1,000 genes with the largest variance across all cells. The gene selection
uses every cell, so it is a global choice for display, not a held-out one. The
factors below are fitted here with `fit_ebmf()` using non-negative
point-exponential priors; they are not the published Jamboree factors.

```{r fit}
pancreas_dir <- Sys.getenv("FLASHIER_UTILS_PANCREAS_DIR")
cache <- readRDS(file.path(pancreas_dir, "pancreas-cells.rds"))
metadata <- cache$metadata
metadata$sample <- as.character(metadata$id)
rownames(metadata) <- metadata$sample
metadata$celltype <- droplevels(factor(metadata$celltype))
metadata$tech <- droplevels(factor(metadata$tech))
counts <- cache$counts
stopifnot(identical(rownames(counts), metadata$sample))
Y <- Diagonal(x = cache$reference_library_mean / rowSums(counts)) %*% counts
Y <- log1p(Y)
dimnames(Y) <- dimnames(counts)
variance <- colMeans(Y^2) - colMeans(Y)^2
genes <- names(sort(variance, decreasing = TRUE))[1:1000]
X <- as.matrix(Y[, genes])
fit <- fit_ebmf(X, sample_side = "rows", ebnm_fn = ebnm::ebnm_point_exponential,
                max_factors = 8, backfit = TRUE, nullcheck = FALSE, seed = 1,
                verbose = 0,
                feature_scale = "shifted_log_reference_library_normalized_measurements")
view <- standardize_factors(fit)
dim(view$activity)
```

Mean activity by group, scaled so the largest group is 1, shows what each factor
tracks.

```{r group-means}
group_means <- function(by) {
  m <- rowsum(view$activity, metadata[[by]]) / as.vector(table(metadata[[by]]))
  round(sweep(m, 2, apply(m, 2, max), "/"), 2)
}
group_means("celltype")
group_means("tech")
```

Four factors follow cell types and four follow technologies. `F2` is acinar and
`F3` is ductal. `F4` is alpha, with gamma at about 0.6 of its maximum, and `F5`
is beta, with delta at about 0.8. `F1`, `F6`, `F7` and `F8` are each largest in a
single technology (CEL-Seq2, Fluidigm C1, inDrop3 and SMART-Seq2, respectively),
so they encode platform differences that remain after library-size scaling.
Group means weight cells equally and the Fluidigm C1 groups are smaller, so
a column such as `F6` can also show cell-type differences that come from that
technology's composition.

## Add activities and compute an embedding

Use raw counts for the same 1,000 genes as the Seurat assay, so feature
identifiers match the factor effects and loadings are stored as well.

```{r seurat}
if (requireNamespace("SeuratObject", quietly = TRUE) && requireNamespace("uwot", quietly = TRUE)) {
  object <- SeuratObject::CreateSeuratObject(
    t(counts[, genes]), meta.data = metadata[, c("tech", "celltype")])
  object <- add_factors_to_seurat(object, view)
  object <- embed_factors(object, seed = 15, n_neighbors = 15, n_epochs = 100)
  plot_factor_embedding(object, "ebmfumap", paste0("F", 1:8))
}
```

Replace the object with your existing Seurat object. Its cell IDs must exactly
match the activity sample IDs. Feature loadings are included only when assay
feature IDs match. Compare the embedding against the metadata it did not use:

```{r embedding-metadata}
if (requireNamespace("SeuratObject", quietly = TRUE) && requireNamespace("uwot", quietly = TRUE)) {
  umap <- as.data.frame(SeuratObject::Embeddings(object[["ebmfumap"]]))
  colnames(umap) <- c("UMAP1", "UMAP2")
  umap$celltype <- object$celltype
  umap$tech <- object$tech
  ggplot2::ggplot(umap, ggplot2::aes(UMAP1, UMAP2, color = celltype, shape = tech)) +
    ggplot2::geom_point(size = 1)
}
```

In the embedding, each combination of technology and cell type forms its own
compact group, and the same cell type from different technologies is not merged.
This is expected because the four technology factors are part of the coordinates.
UMAP visualizes factor-space geometry. It does not perform batch correction or
establish replication.

## Display activity percentiles

A percentile expresses activity relative to a declared set of reference cells.
Here the reference is the 240 CEL-Seq2 cells, so
the percentile of a cell from another technology is partly a technology effect.

```{r percentiles}
reference <- rownames(metadata)[metadata$tech == "celseq2"]
percentile <- factor_activity_percentile(view, reference_samples = reference)
head(percentile)
```

```{r percentile-embedding}
if (requireNamespace("SeuratObject", quietly = TRUE) && requireNamespace("uwot", quietly = TRUE)) {
  plot_factor_embedding(object, "ebmfumap", c("F4", "F5"), display = "percentile",
                        reference_samples = reference)
}
```

Retain the original activities for association tests. The cache
carries no donor identifiers, so cells are not independent units and donor-level
inference is not possible here. See [Compare groups and test associations](contrasts.html).

## Sources

The data are from Peter Carbonetto's
[single-cell Jamboree](https://github.com/stephenslab/single-cell-jamboree)
(Muraro et al., 2016; Luecken et al., 2022, as assembled there).
The reduction and display workflows credit David Zemmour and the Zemmour Lab.
SeuratObject is maintained by Paul Hoffman and the Seurat contributors.
uwot is by James Melville and contributors.
[Original authors and sources](credits.html) lists full authors and citations.

## Function credits

These are the functions from this package that the article calls, with the people
who wrote the upstream code or method behind each one. Use is this package's
classification and code reuse for the function; `provenance()` returns the full
registry. [Original authors and sources](credits.html) gives the full credits,
citations and licenses.

```{r function-credits, echo=FALSE, eval=TRUE}
code <- knitr::knit_code$get()
used <- unique(unlist(lapply(code[names(code) != "function-credits"], function(x)
  tryCatch(all.names(parse(text = x)), error = function(e) NULL))))
registry <- provenance()
credits <- registry[registry[["function"]] %in% used, ]
credits$use <- paste(credits$classification, credits$code_reuse, sep = " / ")
knitr::kable(credits[, c("function", "authors", "conceptual_origin", "use")],
             col.names = c("Function", "Original authors", "Upstream", "Use"),
             row.names = FALSE)
```
