---
title: "Display factors in Seurat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Display factors in Seurat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Store factor activities as a Seurat reduction and plot an embedding
with uniform manifold approximation and projection (UMAP).
Install SeuratObject and uwot to run this example.
Each simulated row represents a cell. Both objects use the same cell IDs.

```{r setup, echo=FALSE}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48), 24, 2), matrix(rnorm(32), 16, 2)) +
  matrix(rnorm(384, sd = 0.2), 24, 16)
dimnames(X) <- list(paste0("s", 1:24), paste0("g", 1:16))
fit <- fit_ebmf(X, sample_side = "rows", seed = 1, max_factors = 2,
                backfit = TRUE, nullcheck = FALSE, verbose = 0,
                feature_scale = "simulated_intensity")
view <- standardize_factors(fit)
```

## Add activities and compute an embedding

```{r seurat}
if (requireNamespace("SeuratObject", quietly = TRUE) && requireNamespace("uwot", quietly = TRUE)) {
  counts <- Matrix::Matrix(
    matrix(1, 16, 24, dimnames = list(colnames(X), rownames(X))), sparse = TRUE
  )
  object <- SeuratObject::CreateSeuratObject(counts)
  object <- add_factors_to_seurat(object, view)
  object <- embed_factors(object, seed = 15, n_neighbors = 4,
                          n_epochs = 20, init = "random")
  plot_factor_embedding(object, "ebmfumap", "F1")
}
```

Replace the placeholder object with your existing Seurat object.
Its cell IDs must exactly match the activity sample IDs.
Feature loadings are included only when assay feature IDs match.
UMAP visualizes factor-space geometry. It does not perform batch correction
or establish replication.

## Display activity percentiles

A percentile expresses activity relative to a declared set of reference cells.

```{r percentiles}
percentile <- factor_activity_percentile(view, reference_samples = rownames(X)[1:12])
head(percentile)
```

Use `display = "percentile"` in `plot_factor_embedding()` to color an embedding
by percentiles. Retain the original activities for association tests.
Donor-level inference still requires a model that accounts for cells sharing
a donor. See [Compare groups and test associations](contrasts.html).

## Sources

The reduction and display workflows credit David Zemmour and the Zemmour Lab.
SeuratObject is maintained by Paul Hoffman and the Seurat contributors.
uwot is by James Melville and contributors.
[Original authors and sources](credits.html) lists full authors and citations.
