---
title: "Check factor stability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Check factor stability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Refit resampled units to assess factor recovery, or compare feature programs
from separate fits. This example uses the GTEx z-score matrix that ships with flashier (`data("gtex", package = "flashier")` ([documentation](https://willwerscheid.github.io/flashier/reference/gtex.html))). Install psych and clue for factor matching.

The tissues in this matrix are not independent donors, so the example does not
resample donors. It places the 1000 SNP-gene pairs in rows, so a factor is a
program across the 44 tissues, and resamples SNP-gene pairs.
Each pair comes from a different gene, but nearby or co-regulated genes need
not be independent. The result describes sensitivity to which pairs were
sampled. It is not donor-level replication. A donor-resampling analysis needs
data with repeated samples from independent donors.

```{r setup}
library(flashier.utils)
data("gtex", package = "flashier")
fit <- fit_ebmf(gtex, sample_side = "rows", max_factors = 3,
                backfit = TRUE, nullcheck = FALSE, seed = 1, verbose = 0,
                feature_scale = "association_z_score")
view <- standardize_factors(fit)
```

```{r metadata}
metadata <- data.frame(sample = rownames(gtex),
                       gene = sub("\\..*$", "", rownames(gtex)))
anyDuplicated(metadata$gene)
```

The gene column is the Ensembl gene ID taken from each row name. The zero shows
that no gene appears twice, so each unit contributes one row.

## Resample units

```{r bootstrap}
if (requireNamespace("psych", quietly = TRUE) && requireNamespace("clue", quietly = TRUE)) {
  stability <- factor_stability(gtex, fit, metadata, unit_col = "gene",
                                B = 5, min_congruence = 0.8, seed = 26)
  print(stability$summary[, c("factor", "n_requested", "n_successful",
                              "recovery_rate_all_requested",
                              "median_congruence", "native_pve_mean")])
  print(stability$failures)
}
```

`B = 5` shows how to run the function. Use more resamples to estimate recovery
in an analysis. The bootstrap resamples whole units with replacement, and F1
and F2 were recovered in every resample here. F3 explains far less variance
(`native_pve_mean`) and was recovered in only some resamples,
although its congruence was high when it was matched.
Choose `min_congruence` for your question. The appropriate matching threshold depends on the data and purpose.
Report failed refits alongside recovery rates.

## Match separate fits

Fit two disjoint random halves of the SNP-gene pairs. Both halves measure the
same 44 tissues, so their tissue programs can be compared directly.

```{r matching}
if (requireNamespace("psych", quietly = TRUE) && requireNamespace("clue", quietly = TRUE)) {
  set.seed(2)
  half <- sample(nrow(gtex), nrow(gtex) / 2)
  fit_half <- function(rows) {
    fit_ebmf(gtex[rows, ], sample_side = "rows", seed = 1, max_factors = 3,
             verbose = 0, backfit = TRUE, nullcheck = FALSE,
             feature_scale = "association_z_score")
  }
  factor_replicability(fit_half(half), fit_half(-half),
                       min_congruence = 0.8)$matches
}
```

Factor numbers need not correspond across fits. Matching compares feature
programs and can account for sign reversals in signed factors.
Here, F1 and F2 match across the two halves, with a lower congruence for F2,
and F3 is not matched. The halves are random subsets of one dataset.
To assess replication, compare independently
collected data with compatible features and preprocessing.
Similar programs alone do not establish that individual activity scores transfer
between cohorts. Use [held-out prediction](holdout.html) to assess prediction.

## Sources

The GTEx data ship with flashier (Jason Willwerscheid, Peter Carbonetto, Wei
Wang and Matthew Stephens) and derive from the GTEx project; see `?flashier::gtex`.

Refits use flashier by Jason Willwerscheid, Peter Carbonetto, Wei Wang and
Matthew Stephens. Matching uses William Revelle's psych and Kurt Hornik and
Walter Böhm's clue. [Original authors and sources](credits.html) lists citations.

## Function credits

These are the functions from this package that the article calls, with the people
who wrote the upstream code or method behind each one. Use is this package's
classification and code reuse for the function; `provenance()` returns the full
registry. [Original authors and sources](credits.html) gives the full credits,
citations and licenses.

```{r function-credits, echo=FALSE, eval=TRUE}
code <- knitr::knit_code$get()
used <- unique(unlist(lapply(code[names(code) != "function-credits"], function(x)
  tryCatch(all.names(parse(text = x)), error = function(e) NULL))))
registry <- provenance()
credits <- registry[registry[["function"]] %in% used, ]
credits$use <- paste(credits$classification, credits$code_reuse, sep = " / ")
knitr::kable(credits[, c("function", "authors", "conceptual_origin", "use")],
             col.names = c("Function", "Original authors", "Upstream", "Use"),
             row.names = FALSE)
```
