---
title: "Check factor stability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Check factor stability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Refit resampled donors to assess factor recovery, or compare feature programs
from separate fits. This example uses simulated data with two visits per donor.
Install psych and clue for factor matching.

```{r setup, echo=FALSE}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48), 24, 2), matrix(rnorm(32), 16, 2)) +
  matrix(rnorm(384, sd = 0.2), 24, 16)
dimnames(X) <- list(paste0("s", 1:24), paste0("g", 1:16))
fit <- fit_ebmf(X, sample_side = "rows", seed = 1, max_factors = 2,
                backfit = TRUE, nullcheck = FALSE, verbose = 0,
                feature_scale = "simulated_intensity")
view <- standardize_factors(fit)
```

```{r metadata}
metadata <- data.frame(sample = rownames(X),
                       donor = rep(paste0("d", 1:12), each = 2))
```

## Resample donors

```{r bootstrap}
if (requireNamespace("psych", quietly = TRUE) && requireNamespace("clue", quietly = TRUE)) {
  stability <- factor_stability(X, fit, metadata, unit_col = "donor",
                                B = 2, min_congruence = 0.8, seed = 26)
  stability$summary[, c("factor", "n_requested", "n_successful",
                         "recovery_rate_all_requested", "median_congruence")]
  stability$failures
}
```

The bootstrap keeps each donor's visits together. `B = 2` shows how to run the
function. Use more resamples to estimate recovery in an analysis.
Choose `min_congruence` for your question. The appropriate matching threshold depends on the data and purpose.
Report failed refits alongside recovery rates.

## Match separate fits

```{r matching}
if (requireNamespace("psych", quietly = TRUE) && requireNamespace("clue", quietly = TRUE)) {
  other <- fit_ebmf(X, sample_side = "rows", seed = 5, max_factors = 2,
                    verbose = 0, backfit = TRUE, nullcheck = FALSE,
                    feature_scale = "simulated_intensity")
  factor_replicability(fit, other, min_congruence = 0.8)$matches
}
```

Factor numbers need not correspond across fits. Matching compares feature
programs and can account for sign reversals in signed factors.
Here, both fits use the same data. To assess replication, compare independently
collected data with compatible features and preprocessing.
Similar programs alone do not establish that individual activity scores transfer
between cohorts. Use [held-out prediction](holdout.html) to assess prediction.

## Sources

Refits use flashier by Jason Willwerscheid, Peter Carbonetto, Wei Wang and
Matthew Stephens. Matching uses William Revelle's psych and Kurt Hornik and
Walter Böhm's clue. [Original authors and sources](credits.html) lists citations.
