---
title: "Understand uncertainty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understand uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Factor uncertainty and regression uncertainty answer different questions.
Factor posterior summaries condition on the learned empirical Bayes priors.
Regression summaries condition on the estimated factor activities.

## Factor posterior intervals

This simulated fit uses priors supported by flashier's posterior sampler.

```{r fit, echo=FALSE}
library(flashier.utils)
set.seed(4)
X <- tcrossprod(matrix(rnorm(40), 20, 2), matrix(rnorm(12), 6, 2)) +
  matrix(rnorm(120, sd = 0.1), 20, 6)
dimnames(X) <- list(paste0("s", 1:20), paste0("f", 1:6))
fit <- fit_ebmf(X, sample_side = "rows", max_factors = 2, seed = 1,
                nullcheck = FALSE, backfit = TRUE, verbose = 0,
                feature_scale = "simulated_intensity",
                ebnm_fn = list(ebnm::ebnm_normal, ebnm::ebnm_point_laplace))
view <- standardize_factors(fit)
```

```{r intervals}
uncertainty <- factor_uncertainty(fit, "activity", factors = "F1", ids = "s1")
uncertainty[, c("sample", "factor", "estimate", "posterior.sd")]

intervals <- factor_intervals(fit, "activity", factors = "F1", ids = "s1",
                              nsamp = 200, seed = 1)
intervals[, c("sample", "factor", "lower", "upper")]
```

The posterior standard deviation is not a regression standard error.
Intervals use native variational draws. Increasing `nsamp` reduces Monte Carlo
variation in the reported quantiles.

## Project coefficient uncertainty

If a model supplies joint draws of factor coefficients, use `input = "draws"`.
Name the columns by factor and supply the activity model's representation.
The random draws below illustrate the required shape.

```{r draws}
set.seed(1)
draws <- matrix(rnorm(200 * ncol(view$activity)), nrow = 200,
                dimnames = list(NULL, colnames(view$activity)))
projected <- backproject_contrast(view, draws, input = "draws",
                                  factor_basis = view$manifest)
head(projected$effects)
```

These intervals hold feature effects fixed. Combining arbitrary factor and
coefficient draws does not create a joint posterior.
For covariance-based projection, supply the full named factor covariance matrix.
Marginal standard errors alone do not recover correlations between coefficients.
Use a separately designed pipeline bootstrap to assess uncertainty from refitting.

## Sources

Posterior sampling uses flashier by Jason Willwerscheid, Peter Carbonetto,
Wei Wang and Matthew Stephens. Reconstruction credits David Zemmour's workflows.
[Original authors and sources](credits.html) lists method citations.
