---
title: "Understand uncertainty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understand uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Factor uncertainty and regression uncertainty answer different questions.
Factor posterior summaries condition on the learned empirical Bayes priors.
Regression summaries condition on the estimated factor activities.

## Factor posterior intervals

This fit uses the GTEx z-score matrix that ships with flashier (`data("gtex", package = "flashier")` ([documentation](https://willwerscheid.github.io/flashier/reference/gtex.html))) and priors
supported by flashier's posterior sampler: a point-Laplace prior for the
SNP-gene pairs and a normal prior for the tissues.

```{r fit}
library(flashier.utils)
data("gtex", package = "flashier")
fit <- fit_ebmf(gtex, sample_side = "columns", max_factors = 2, seed = 1,
                nullcheck = FALSE, backfit = TRUE, verbose = 0,
                feature_scale = "association_z_score",
                ebnm_fn = list(ebnm::ebnm_point_laplace, ebnm::ebnm_normal))
view <- standardize_factors(fit)
```

```{r intervals}
tissues <- c("Whole_Blood", "Brain_Cortex")
uncertainty <- factor_uncertainty(fit, "activity", factors = "F1", ids = tissues)
uncertainty[, c("sample", "factor", "estimate", "posterior.sd")]

intervals <- factor_intervals(fit, "activity", factors = "F1", ids = tissues,
                              nsamp = 200, seed = 1)
intervals[, c("sample", "factor", "lower", "upper")]
```

The posterior standard deviation is not a regression standard error.
Here it is the same for both tissues, which follows from the common normal
prior on tissue activity, and the intervals are narrow relative to the
estimates. Real data have no known truth, so these intervals show the fit's
posterior uncertainty and not tested coverage.
Intervals use native variational draws. Increasing `nsamp` reduces Monte Carlo
variation in the reported quantiles.

## Project coefficient uncertainty

If a model supplies joint draws of factor coefficients, use `input = "draws"`.
Name the columns by factor and supply the activity model's representation.
The random draws below are synthetic. They illustrate the required shape and
are not estimates from a model of the tissues, which are not independent
replicates.

```{r draws}
set.seed(1)
draws <- matrix(rnorm(200 * ncol(view$activity), sd = 0.1), nrow = 200,
                dimnames = list(NULL, colnames(view$activity)))
projected <- backproject_contrast(view, draws, input = "draws",
                                  factor_basis = view$manifest)
head(projected$effects)
```

These intervals hold feature effects fixed. Combining arbitrary factor and
coefficient draws does not create a joint posterior.
For covariance-based projection, supply the full named factor covariance matrix.
Marginal standard errors alone do not recover correlations between coefficients.
Use a separately designed pipeline bootstrap to assess uncertainty from refitting.

## Sources

The GTEx data ship with flashier (Jason Willwerscheid, Peter Carbonetto, Wei
Wang and Matthew Stephens) and derive from the GTEx project; see `?flashier::gtex`.

Posterior sampling uses flashier by Jason Willwerscheid, Peter Carbonetto,
Wei Wang and Matthew Stephens. Reconstruction credits David Zemmour's workflows.
[Original authors and sources](credits.html) lists method citations.

## Function credits

These are the functions from this package that the article calls, with the people
who wrote the upstream code or method behind each one. Use is this package's
classification and code reuse for the function; `provenance()` returns the full
registry. [Original authors and sources](credits.html) gives the full credits,
citations and licenses.

```{r function-credits, echo=FALSE, eval=TRUE}
code <- knitr::knit_code$get()
used <- unique(unlist(lapply(code[names(code) != "function-credits"], function(x)
  tryCatch(all.names(parse(text = x)), error = function(e) NULL))))
registry <- provenance()
credits <- registry[registry[["function"]] %in% used, ]
credits$use <- paste(credits$classification, credits$code_reuse, sep = " / ")
knitr::kable(credits[, c("function", "authors", "conceptual_origin", "use")],
             col.names = c("Function", "Original authors", "Upstream", "Use"),
             row.names = FALSE)
```
