---
title: "Method-specific pathway summaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Method-specific pathway summaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48),24,2),matrix(rnorm(32),16,2)) + matrix(rnorm(384,sd=0.2),24,16)
dimnames(X) <- list(paste0("s",1:24),paste0("g",1:16))
fit <- fit_ebmf(X,"rows",seed=1,max_factors=2,verbose=0,backfit=TRUE,nullcheck=FALSE,feature_scale="arcsinh_intensity")
view <- standardize_factors(fit)
metadata <- data.frame(sample=rownames(X),donor=rep(paste0("d",1:12),each=2),group=rep(c("A","B"),each=12),visit=rep(0:1,12))
```

Ranked fgsea enrichment and decoupleR network scores are different estimands. Feature effects and least-extreme distinctiveness are explicit alternative signals. Annotation universes, sizes and losses are retained.
```{r}
pathways <- list(panelA=paste0("g",1:6),panelB=paste0("g",7:12))
if (requireNamespace("fgsea",quietly=TRUE)) {
  enrichment <- enrich_factors(view,pathways,"fgsea",factors="F1",control=list(seed=12,nPermSimple=100L,eps=1e-3,minSize=3L))
  enrichment$table
}
```
For real gene sets attach a version and source to the named list. These simulated labels are a structural example without biological meaning. SuSiE is explicitly unsupported until its optional source and native API gates are completed.

Signed network weights can annotate factor feature programs directly. The rows
of the input are features and the columns are factors, as required by decoupleR.
Supply a network with `source`, `target`, and `mor` columns; for gene-based
features, native resources such as `decoupleR::get_collectri()` or
`decoupleR::get_progeny()` can provide these networks. PROGENy returns a `weight`
column: set `network$mor <- network$weight` before passing it here. Cache and
version real annotations before use. Feature IDs must match network targets, or be mapped
explicitly with a one-to-one `feature_map`.

```{r}
network <- data.frame(source=rep(c("programA","programB"),each=6),
                      target=paste0("g",1:12),
                      mor=rep(c(1,-1,0.5,-0.5,2,-2),2))
attr(network,"annotation_version") <- "simulated_v1"
if (requireNamespace("decoupleR",quietly=TRUE)) {
  annotation <- enrich_factors(view,network,"decoupleR",
                                control=list(method="ulm",minsize=3L))
  annotation$table
  annotation$analysis_metadata$pathway_sizes
}
```

`condition` identifies the factor in the native result. ULM and MLM return native
scores and p-values. The entire mapped feature universe is passed to the native
method, including features without network edges. Missing targets and mapping
losses are recorded, along with a hash of the signed network. Duplicate edges
are rejected rather than silently summed. Native `minsize` filtering and errors
are retained. Use `control=list(method="mlm",minsize=3L)` for multivariate scoring.

Scores describe the estimated feature program, not per-sample pathway activity.
Reversing a factor's sign reverses its annotation score. By default the input is
not centered across factors; `center=TRUE` changes the signal and makes the
result depend on which factors were selected. Native p-values condition on the
estimated weights and do not propagate EBMF uncertainty. These simulated
programs have no biological meaning. For CyTOF or other non-gene features, use
networks matched to those measured features rather than assuming a gene network
applies to a marker panel.
