---
title: "Held-out projection and masked prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Held-out projection and masked prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48),24,2),matrix(rnorm(32),16,2)) + matrix(rnorm(384,sd=0.2),24,16)
dimnames(X) <- list(paste0("s",1:24),paste0("g",1:16))
fit <- fit_ebmf(X,"rows",seed=1,max_factors=2,verbose=0,backfit=TRUE,nullcheck=FALSE,feature_scale="arcsinh_intensity")
view <- standardize_factors(fit)
metadata <- data.frame(sample=rownames(X),donor=rep(paste0("d",1:12),each=2),group=rep(c("A","B"),each=12),visit=rep(0:1,12))
```

Named folds keep both visits from each donor together. The callback fits programs on training data only. Identity preprocessing is declared explicitly; globally fitted normalization must be labelled conditional-global.
```{r}
folds <- data.frame(sample=metadata$sample,fold=rep(c("training","held"),c(16,8)))
callback <- function(X,metadata,sample_side,seed,settings,observation_aux=list()) {
  fit_ebmf(X,sample_side,seed=seed,max_factors=2,verbose=0,backfit=TRUE,nullcheck=FALSE,feature_scale="arcsinh_intensity")
}
control <- list(sample_side="rows",preprocessing_manifest=list(scope="identity"),evaluated_folds="held")
projection <- factor_holdout(X,metadata,"donor",folds,callback,evaluation="projection",seed=1,control=control)
projection$metrics
masked <- factor_holdout(X,metadata,"donor",folds,callback,evaluation="masked",activity_features=paste0("g",1:8),scoring_features=paste0("g",9:16),seed=1,control=control)
masked$metrics
```
Projection uses observed test entries to estimate activities. Masked scoring excludes the scoring entries from that estimation and from test noise/prior calibration. They are different evaluations. Programs remain fixed; nuisance parameters can recalibrate on permitted test entries. Freezing training noise is unsupported. Leave-one-cohort-out uses cohort IDs as fold labels and donor IDs as units, provided each donor belongs to one cohort.
