---
title: "Largest and distinctive feature effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Largest and distinctive feature effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48),24,2),matrix(rnorm(32),16,2)) + matrix(rnorm(384,sd=0.2),24,16)
dimnames(X) <- list(paste0("s",1:24),paste0("g",1:16))
fit <- fit_ebmf(X,"rows",seed=1,max_factors=2,verbose=0,backfit=TRUE,nullcheck=FALSE,feature_scale="arcsinh_intensity")
view <- standardize_factors(fit)
metadata <- data.frame(sample=rownames(X),donor=rep(paste0("d",1:12),each=2),group=rep(c("A","B"),each=12),visit=rep(0:1,12))
```

Largest absolute effects answer which measurements contribute most. Least-extreme distinctiveness compares each factor with all other factors in the same declared basis. These rankings have different meanings and have no significance test.
```{r}
factor_features(view,"F1",select="largest",n=4)
factor_features(view,"F1",select="distinctive",n=4)
plot_factor_features(view,"F1",select="largest",n=4)
```
The two ranking helpers are licensed adaptations of singlecelljamboreeR, not new statistical estimators. Ties are deterministic by feature ID for display.
