---
title: "Comparable multi-condition effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparable multi-condition effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48),24,2),matrix(rnorm(32),16,2)) + matrix(rnorm(384,sd=0.2),24,16)
dimnames(X) <- list(paste0("s",1:24),paste0("g",1:16))
fit <- fit_ebmf(X,"rows",seed=1,max_factors=2,verbose=0,backfit=TRUE,nullcheck=FALSE,feature_scale="arcsinh_intensity")
view <- standardize_factors(fit)
metadata <- data.frame(sample=rownames(X),donor=rep(paste0("d",1:12),each=2),group=rep(c("A","B"),each=12),visit=rep(0:1,12))
```

All condition coefficients must use one activity basis, one comparable term and all eligible factor rows. Here a simulated coefficient table demonstrates native mash execution. This does not establish empirical Bayes calibration with two factors.
```{r}
ids <- view$manifest$factor_ids
conditions <- c("cohortA","cohortB")
table <- data.frame(factor=rep(ids,2),condition=rep(conditions,each=length(ids)),term="B-minus-A",estimate=seq_len(2*length(ids))/5,std.error=0.5,basis_id=view$manifest$basis_id)
attr(table,"analysis_metadata") <- view$manifest
if (requireNamespace("mashr",quietly=TRUE)) {
  result <- shrink_factor_effects(table,"condition",term="B-minus-A",independent_conditions=TRUE)
  result$table
}
factor_cor(view,"pearson")$correlation
```
Supply V when condition estimation errors are correlated; identity V requires an explicit independence declaration. Few and dependent factor rows limit covariance learning. Pooled change, heterogeneity and sharing are distinct scientific claims. Correlations are descriptive and ignore posterior score uncertainty. Optional brms and uncertain-response modes are deferred; native Gaussian `se(...,sigma=TRUE)` syntax was inspected, but no posterior model was fitted or calibrated.
