---
title: "Representation and measurement units"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Representation and measurement units}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48),24,2),matrix(rnorm(32),16,2)) + matrix(rnorm(384,sd=0.2),24,16)
dimnames(X) <- list(paste0("s",1:24),paste0("g",1:16))
fit <- fit_ebmf(X,"rows",seed=1,max_factors=2,verbose=0,backfit=TRUE,nullcheck=FALSE,feature_scale="arcsinh_intensity")
view <- standardize_factors(fit)
metadata <- data.frame(sample=rownames(X),donor=rep(paste0("d",1:12),each=2),group=rep(c("A","B"),each=12),visit=rep(0:1,12))
```

D is absorbed exactly once. Both canonical coordinates reconstruct the same native posterior mean.
```{r}
raw <- standardize_factors(fit,scaling="raw")
ldf <- standardize_factors(fit,d_location="activity")
stopifnot(isTRUE(all.equal(tcrossprod(raw$activity,raw$effects),tcrossprod(ldf$activity,ldf$effects),tolerance=1e-10)))
ldf$manifest[c("basis_id","feature_scale","d_location","orientation")]
```
Changing normalization changes the coefficient basis. For a baseline-standardization recipe, compute means and SD on the declared baseline reference samples, transform activity, divide its matched coefficient scale accordingly and reconstruct with effects multiplied by the reference SD. A centered basis also requires an explicit feature offset. This package does not silently create that transformed basis or transport external coefficients into it.
