---
title: "Donor stability and independent-fit replicability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Donor stability and independent-fit replicability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(flashier.utils)
set.seed(3)
X <- tcrossprod(matrix(rnorm(48),24,2),matrix(rnorm(32),16,2)) + matrix(rnorm(384,sd=0.2),24,16)
dimnames(X) <- list(paste0("s",1:24),paste0("g",1:16))
fit <- fit_ebmf(X,"rows",seed=1,max_factors=2,verbose=0,backfit=TRUE,nullcheck=FALSE,feature_scale="arcsinh_intensity")
view <- standardize_factors(fit)
metadata <- data.frame(sample=rownames(X),donor=rep(paste0("d",1:12),each=2),group=rep(c("A","B"),each=12),visit=rep(0:1,12))
```

The bootstrap draws donors and preserves both visits, with unique occurrence IDs. Threshold-free assignments are candidates. A caller-chosen congruence criterion gives accepted matches without a universal stable/unstable label.
```{r}
if (requireNamespace("psych",quietly=TRUE) && requireNamespace("clue",quietly=TRUE)) {
  stability <- factor_stability(X,fit,metadata,"donor",B=2,min_congruence=0.8,seed=26)
  stability$summary
  stability$failures
  other <- fit_ebmf(X,"rows",seed=5,max_factors=2,verbose=0,backfit=TRUE,nullcheck=FALSE,feature_scale="arcsinh_intensity")
  factor_replicability(fit,other,min_congruence=0.8)$matches
}
```
This tiny bootstrap demonstrates execution; it cannot estimate recovery reliably. Full fixed-budget benchmark results live outside the vignette. Report both successful-fit and all-requested denominators. Concordance between independently fitted programs does not establish measurement invariance, individual-level transportability or biological replication by itself.
